---
title: "Analysis of iPSC Gene Expression (will be changed)"
subtitle: "Final Project"
author: "Trinity Minard"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
format:
  pdf:
    keep-tex: true
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: inline
bibliography:
  - references.bib
  - packages.bib
csl: norsk-apa-manual.csl
---

# Introduction

The discovery of human induced pluripotent stem cells (iPSCs) by
Takahashi and Yamanaka in 2007 [@takahashi2007] has revolutionized the
regenerative medicine field. iPSCs are created by reprogramming somatic
cells into an embryonic like state using the Yamanaka transciption
factors: SOX2, OCT4, KLF4, and c-Myc). These cells have vast
capabilities due to their ability to model human development and
diseases *in vitro*, and their potential for patient specific
(autologous) and "off the shelf" (allogeneic) cell therapies
[@cerneckis2024].

With that being said, the characterization of iPSC lines is essential in
any experimental or manufacturing process to ensure consistency in
culturing. This includes analyzing the expression of pluripotency (stem
cell) markers, differentiation markers, and potential chromosomal
abnormalies, to access the population of the culture. This is commonly
done using techniques such as reverse transcriptase or real time
quantitative PCR (RT-qPCR) to analyze mRNA expression [@artyukhov2017],
or using an antibody based experiment to analyze protein expression
levels, such as Flow Cytometry or immunocytochemistry. When conducting
these experiments, specifically RT-qPCR, reference genes, also known as
housekeeping genes, are used to normalize the expression of other genes
in a sample [@artyukhov2017]. This allows the expression of genes to be
compared across different samples, since it is presumed that the
expression of the housekeeping gene is consistent across all of the
samples. In stem cell samples, however, it has been found that the
commonly used housekeeping genes, such as GAPDH, appear to fluctuate in
iPSC samples [@panina2020], making it difficult to achieve accurate
normalization data.

RNA Seq is a technique that sequences the entirety of RNA molecules in a
given sample, rather than using gene specific primers or probes as in
RT-qPCR and other techniques. This allows the entire transciptome to be
analyzed, providing a complete picture of the expression of genes in a
sample [@wang2009]. Additionally, gene expression is calculated by
dividing the raw counts of a gene by the total counts in a sample,
eliminating the need for a housekeeping gene. This allows for an
unbiased and quantitiative analysis of millions of genes in sample
[@wang2009]. Despite these advantages, RNA-Seq is far more time
consuming and expensive than RT-qPCR and other experiments, making it
non practical in a manufacturing setting. Therefore, utilizing RNA-Seq
to analyze iPSC samples could provide insight into housekeeping gene
expression, allowing researchers to select ideal reference genes for
their analysis.

A similar study, conducted by Carcamo-Orive *et al.,* investigated the
variability of gene expression across 317 iPSC lines generated from 101
individuals [@carcamo-orive2017] and created a publically avaible data
set of their raw RNA Seq data. They found that genetic variability in
the iPSC lines was influenced by a donor specific gene expression
pattern, specifically influenced by donor age, sex, ancestry, and
reprogramming variabilies. iPSCs can also be reprogrammed from a variety
of somatic cells, such as blood, skin fibroblasts, hair keratinocytes,
urine cells, mesencymal stem cells, and bone marrow [@poetsch2022].
While some somatic cells are easier to obtain than others, such as blood
cells and fibroblasts, iPSCs from these lines are associated with higher
genomic mutations due to the somatic cells' high turnover rate.
Additionally, iPSCs appear to have an epigenetic memory of their tissue
of origin, which has been found to influence their differentiation
[@poetsch2022]. All of these factors, and many others, could be
contributing to the variations in gene expression in iPSC samples.

This publicly available dataset from Carcamo-Orive *et al.* was used to
analyze the expression of RNA across iPSC samples, with a focus on the
expression of pluripotency markers, differentiation markers, and
housekeeping genes.

# Methods

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#create list of packages to cite
packages <- c("base", "tidyverse", "dplyr", "ggplot2", "scales", "pheatmap", "lme4", "lmerTest", "emmeans", "broom", "ggsignif", "ggsci", "patchwork", "FactoMineR", "grid", "factoextra", "rstatix", "kableExtra")

#citing R and packages
knitr::write_bib(packages, "packages.bib")
```

### Data Cleaning

RNA-Seq counts per million data was acquired from Stemformatics using
the Carcamo-Orive (2017) dataset [@carcamo-orive2017]. Genes in this
study were referred to by their Ensembl ID, so gene information, such as
name and RNA type, was integrated from the Emsembl database.

*further cleaning of the count per million data will be added*

### Figure 1

Genes detected in the RNA Seq counts per million data were grouped into
the following catagories: Protein coding, long non coding RNA (lncRNA),
Small RNAs, Pseudogenes, and Other. A count of the number of genes in
each category was then performed by using the dplyr package [@R-dplyr]
from tidyverse [@R-tidyverse]. A fraction and percentage were then
calculated by dividing the count by the sum of count (i.e. total count
for all categories). The scales [@R-scales] package from ggplot2
[@R-ggplot2] was used to create a pie chart of the counts by RNA
category.

*continue methods for following figures ...*

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#library packages
library(tidyverse)
library(dplyr)
library(scales)
library(ggrepel)
library(pheatmap)
library(lme4)
library(lmerTest)  
library(emmeans)
library(broom)
library(ggplot2)
library(ggsignif)
library(ggsci)
library(patchwork)
library(FactoMineR)
library(grid) 
library(factoextra)
library(rstatix)
library(kableExtra)
library(lattice)

#import data files
raw_counts <- read_tsv("stemformatics_dataset_7274.samples.tsv")
cpm_data   <- read_tsv("stemformatics_dataset_7274.cpm.tsv")
ensembl_data <- read_csv("mart_export.txt")

# view data files
#head(raw_counts)
#head(cpm_data)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#Change column name from "0.001" to "Gene Stable ID"
cpm_data <- cpm_data %>%
  rename("Gene stable ID" = "...1")

#Create a data frame with only the gene info for the genes in this study
ensembl_filtered <- ensembl_data %>%
  filter(`Gene stable ID` %in% cpm_data$`Gene stable ID`)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Add a new column 'Category' based on 'Gene type'
ensembl_filtered <- ensembl_filtered %>%
  mutate(Category = case_when(
    `Gene type` %in% c("protein_coding") ~ "Protein coding",
    `Gene type` %in% c("lncRNA", "TEC") ~ "lncRNA",
    `Gene type` %in% c("miRNA", 
                          "snoRNA", 
                          "snRNA", 
                          "scaRNA", 
                          "sRNA", 
                          "vault_RNA", 
                          "ribozyme") ~ "Small RNAs",
    `Gene type` %in% c("pseudogene", 
                          "processed_pseudogene", 
                          "unprocessed_pseudogene",
                          "transcribed_unprocessed_pseudogene",
                          "transcribed_processed_pseudogene",
                          "unitary_pseudogene", 
                          "IG_C_pseudogene", 
                          "IG_V_pseudogene",
                          "IG_J_pseudogene", 
                          "TR_J_pseudogene", 
                          "TR_V_pseudogene",
                          "translated_processed_pseudogene",
                        "transcribed_unitary_pseudogene") ~ "Pseudogenes",
       `Gene type` %in% c("misc_RNA", 
                          "artifact", 
                          "Mt_tRNA", 
                          "IG_V_gene", 
                          "IG_J_gene", 
                          "IG_C_gene", 
                          "IG_D_gene",
                          "TR_J_gene", 
                          "TR_V_gene", 
                          "TR_D_gene", 
                          "TR_C_gene",
                          "rRNA", 
                          "Mt_rRNA", 
                          "rRNA_pseudogene") ~ "Other",
   TRUE ~ "Other"))

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#summarize count of genes by RNA category
RNA_types_summary <- ensembl_filtered %>%
  count(Category) %>%  # creates a numeric column `n`
  mutate(
    fraction   = n / sum(n),
    label_text = paste0(Category, "\n", round(fraction * 100, 1), "%"),
    large_slice = fraction > 0.1)

# pivot raw cpm data and add ensembl data information by Gene. then remove NAs
cpm_long <- cpm_data %>%
  pivot_longer(-`Gene stable ID`, names_to = "Sample", values_to = "Expression") %>%
  left_join(ensembl_filtered, by = "Gene stable ID") %>%
  filter(!is.na(Category), !is.na(Expression)) 

# summarize expression per category 
cpm_category_summary <- cpm_long %>%
  group_by(Category) %>%
  summarize(Total_Expression = sum(Expression)) %>%
  ungroup() %>%
  mutate(
    fraction = Total_Expression / sum(Total_Expression),
    label_text = paste0(Category, "\n", round(fraction * 100, 1), "%"),
    label_pos = cumsum(fraction) - fraction / 2,
    large_slice = fraction > 0.1) 
```

# Results

```{r, echo=FALSE, warning=FALSE, message=FALSE}

#manually set colors for each category (using factors was causing issues when plotting)
slice_colors <- c(
  "Protein coding" = "#7570b3",
  "Pseudogenes"    = "#d95f02",
  "lncRNA"         = "#1b9e77",
  "Small RNAs"     = "#e7298a",
  "Other"          = "#66a61e")

#pie chart of RNA categories 
plot_RNA_types <- ggplot(RNA_types_summary, aes(x = "", y = fraction, fill = Category)) +
  geom_bar(stat="identity", width=1, color="white") +
  scale_fill_manual(values = slice_colors) +
  coord_polar(theta="y") +
  theme_void() +
  theme(
    legend.position = "none",
    plot.margin = margin(0,0,0,0))

# Extract actual slice positions
pb <- ggplot_build(plot_RNA_types)
bar_positions <- pb$data[[1]]  # first layer = geom_bar
bar_positions$Category <- RNA_types_summary$Category
bar_positions$label_text <- RNA_types_summary$label_text
bar_positions$large_slice <- RNA_types_summary$large_slice

# Compute label positions for all slices
bar_positions <- bar_positions %>%
  mutate(label_pos = (ymin + ymax)/2)

# Add labels
plot_RNA_types_final <- plot_RNA_types +
  # On-slice labels for large slices
  geom_text(
    data = subset(bar_positions, large_slice),
    aes(y = label_pos, label = label_text),
    color = "white",
    fontface = "bold",
    size = 4) +
  # Repelled labels for small slices
  geom_text_repel(
    data = subset(bar_positions, !large_slice),
    aes(x = 1, y = label_pos, label = label_text),
    nudge_x = 0.8,
    force = 0.5,
    segment.color = "grey60",
    segment.length = 0.2,
    size = 4,
    fontface = "bold",
    inherit.aes = FALSE) +
  labs(caption = "Number of Genes per RNA Category") +
  theme(plot.caption = element_text(size=12, hjust=0.5, face="bold"))

#pie chart of RNA expression
plot_expression <- ggplot(cpm_category_summary, aes(x = "", y = fraction, fill = Category)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  scale_fill_manual(values = slice_colors) +
  coord_polar(theta = "y") +
  theme_void() +
  theme(
    legend.position = "none",
    plot.margin = margin(0, 0, 0, 0))

# Extract actual slice positions from ggplot
pb <- ggplot_build(plot_expression)
bar_positions <- pb$data[[1]]  # first layer = geom_bar
bar_positions$Category <- cpm_category_summary$Category
bar_positions$label_text <- cpm_category_summary$label_text
bar_positions$large_slice <- cpm_category_summary$large_slice

# Compute label positions
bar_positions <- bar_positions %>%
  mutate(label_pos = (ymin + ymax)/2)

# Add labels
plot_RNA_expression_final <- plot_expression +
  # On-slice labels for large slices
  geom_text(
    data = subset(bar_positions, large_slice),
    aes(y = label_pos, label = label_text),
    color = "white",
    fontface = "bold",
    size = 4) +
  # Repelled labels for small slices
  geom_text_repel(
    data = subset(bar_positions, !large_slice),
    aes(x = 1, y = label_pos, label = label_text),
    nudge_x = 0.8,
    force = 0.5,
    segment.color = "grey60",
    segment.length = 0.2,
    size = 4,
    fontface = "bold",
    inherit.aes = FALSE) +
  labs(caption = "Proportion of Total Expression per RNA Category") +
  theme(plot.caption = element_text(size = 12, hjust = 0.5, face = "bold"))
```

```{r fig-rna-pie-chart, fig.cap = "Detected RNA Types in iPSC Sample Analzyed using RNA Seq. The number of genes detected in each RNA category from RNA-seq data were counted and plotted as a pie chart (left). Small categories (< 2 %), such as rRNA, tRNA, and immunoglobulin genes, were combined into an 'Other' category. The expression of genes, using raw count per million data, was averaged by type of RNA and plotted as a pie chart (right).", fig.width=10, fig.height=5, echo=FALSE, warning=FALSE, message=FALSE}
plot_RNA_types_final + plot_RNA_expression_final 
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#begin cleaning data 

#verify that the sample ids match in raw_counts and _cpm_data
raw_counts <- raw_counts[match(colnames(cpm_data)[-1], raw_counts$sample_id), ]

#all(colnames(cpm_data)[-1] == raw_counts$sample_id)

#convert counts per million to a log scale, excluding the first column which has gene info, and adding an offset so that values use to 0 are not skewed
logCPM <- cbind(`Gene stable ID` = cpm_data$`Gene stable ID`, log2(cpm_data[,-1] + 0.5))

#summary(raw_counts)
#unique(raw_counts$parental_cell_type) #all parent cells are blood mononuclear
#unique(raw_counts$disease_state) #all normal
#unique(raw_counts$sex) #male and female
#unique(raw_counts$developmental_stage) #33 ages 

#based on information regarding the samples, sex, age and race differ across the samples.

#age is reported as "age years old", so a new column was added with just the numeric
raw_counts$age_numeric <- as.numeric(sub("-year-old.*", "", raw_counts$developmental_stage))

#age was then grouped by decade to reduce the number of conditions 
raw_counts$age_decade <- cut(
  raw_counts$age_numeric,
  breaks = seq(40, 80, by = 10),
  labels = c("40s", "50s", "60s", "70s"),
  right = FALSE)

#create a new factor column in the raw counts data which combined age and sex 
raw_counts_cleaned <- raw_counts %>%
 filter(!is.na(age_decade) & !is.na(sex)) %>%
  mutate(age_sex = factor(paste(age_decade, sex, sep = " "))) %>%
  mutate(
    # iPSC line ID (number_number)
    line_id = str_extract(sample_description, "\\d+_\\d+"),
    
    # sex: look for "female" or "male"
    sex = str_extract(sample_description, "female|male"),
    
    # age in years
    ageYr = as.numeric(str_extract(sample_description, "\\d+(?=yr)")),
    
    # race
    race = str_extract(sample_description, "caucasian|Asian|South Asian|African American|Hispanic"),
    
    # insulin status: "Insulin sensitive" or "Insulin resistant"
    insulin_status = str_extract(sample_description, "Insulin [a-zA-Z]+")) %>%
  mutate(
    sex = factor(sex),
    race = factor(race),
    insulin_status = factor(insulin_status)) %>%
  mutate(race = recode(race,
                       "caucasian" = "Caucasian"))

#select columns in raw counts which have minimal NAs and are variable
raw_counts_cleaned <- raw_counts_cleaned %>%
  select(sample_id, sex, reprogramming_method, 
         external_source_id, age_numeric, age_decade, 
         age_sex, race, insulin_status)

# add gene name from ensembl filtered to longCPM data
logCPM_named <- logCPM %>%
  left_join(
    ensembl_filtered %>% dplyr::select(`Gene stable ID`, `Gene name`),
    by = c("Gene stable ID" = "Gene stable ID"))

# move gene name to first column
logCPM_named <- logCPM_named %>%
  dplyr::select(`Gene name`, everything())
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#view n numbers by category and create kable tables 
raw_counts_cleaned %>% 
  count(age_sex) %>% 
  arrange(age_sex) %>% 
  rename(`Age-Sex Group` = age_sex, `Sample Count` = n) %>% 
  kable("latex", booktabs = TRUE, caption = "Sample counts by Age-Sex") %>% 
  kable_styling(latex_options = c("striped", "hold_position"))

raw_counts_cleaned %>% 
  count(race) %>% 
  arrange(race) %>% 
  rename(`Race` = race, `Sample Count` = n) %>% 
  kable("latex", booktabs = TRUE, caption = "Sample counts by Race") %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#remove rows from log_CPM_named when gene is NA and remove duplicates
logCPM_named_clean <- logCPM_named %>%
  filter(!is.na(`Gene name`) & `Gene name` != "") %>%
  distinct(`Gene name`, .keep_all = TRUE) 

#create a matrix of the expression data excluding gene info 
mat <- logCPM_named_clean %>%
  select(where(is.numeric)) %>%
  as.matrix()

#set each row in the matrix as its corresponding gene name 
rownames(mat) <- logCPM_named_clean$`Gene name`

#calculates gene variance 
gene_var <- apply(mat, 1, var)

#removes genes with minimal variance to reduce clutter in the plot
mat_filtered <- mat[gene_var > 0, ]

#selects the top 20 genes with the most variance to label in the plot
top_n <- 20  # number of genes to show as arrows 

top_genes <- names(sort(gene_var[gene_var > 0], decreasing = TRUE))[1:top_n]

mat_top <- mat_filtered[top_genes, ]

mat_t_top <- t(mat_top)

#turn top genes into a data frame
top_genes_df <- data.frame(
  Gene = top_genes,
  Variance = gene_var[top_genes]) %>%
  arrange(desc(Variance))

#transpose the matrix to have samples as rows and genes as columns for FactoMineR
mat_t <- t(mat_filtered)  

#perform principal component analysis 
pca_res <- PCA(mat_t, graph = FALSE)

#create sample coordinates and add them as columns PC1 and PC2
pca_df <- data.frame(pca_res$ind$coord[,1:2])

colnames(pca_df) <- c("PC1", "PC2")

pca_df$Sample <- rownames(pca_df)

#add sample info from raw_counts to the pca dataframe 
pca_df <- pca_df %>%
  left_join(raw_counts_cleaned, by = c("Sample" = "sample_id"))

#calculate percent variance 
var_explained <- round(pca_res$eig[1:2, "percentage of variance"], 1)

# use anova to see if sex, age, or race influence PC1 or PC2 

#sex alone
#oneway.test(PC1 ~ sex, data = pca_df, var.equal = FALSE)
#oneway.test(PC2 ~ sex, data = pca_df, var.equal = FALSE)

#age alone
#oneway.test(PC1 ~ age_decade, data = pca_df, var.equal = FALSE)
#oneway.test(PC2 ~ age_decade, data = pca_df, var.equal = FALSE)

#age and decade
#oneway.test(PC1 ~ age_sex, data = pca_df, var.equal = FALSE)
#oneway.test(PC2 ~ age_sex, data = pca_df, var.equal = FALSE)

#race alone
#oneway.test(PC1 ~ race, data = pca_df, var.equal = FALSE)
#oneway.test(PC2 ~ race, data = pca_df, var.equal = FALSE)
```

```{r fig-pca-sexage, fig.cap = "Caption.", fig.width=10, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
#plot age_sex
pca_age_sex <- fviz_pca_biplot(
  pca_res,
  geom.ind = "point",                # samples as points
  habillage = pca_df$age_sex,           # color points by race
  addEllipses = TRUE,                # ellipses for race groups
  ellipse.level = 0.95,
  repel = TRUE,                      # avoid overlapping arrows
  geom.var = c("arrow", "text"),     # arrows + gene labels
  select.var = list(name = top_genes),  # only top variable genes
  col.var = "red",                    # arrows in red
  alpha.var = 0.7) +                   # slightly transparent arrows
labs(
  x = paste0("PC1 (", var_explained[1], "%)"),
  y = paste0("PC2 (", var_explained[2], "%)")) +
theme_classic(base_size = 14) +
theme(legend.position = "top")

pca_age_sex
```

```{r fig-pca-race, fig.cap = "Caption.", fig.width=10, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
#plot race
pca_race <- fviz_pca_biplot(
  pca_res,
  geom.ind = "point",                # samples as points
  habillage = pca_df$race,           # color points by race
  addEllipses = TRUE,                # ellipses for race groups
  ellipse.level = 0.95,
  repel = TRUE,                      # avoid overlapping arrows
  geom.var = c("arrow", "text"),     # arrows + gene labels
  select.var = list(name = top_genes),  # only top variable genes
  col.var = "red",                    # arrows in red
  alpha.var = 0.7) +               # slightly transparent arrows 
labs(
  x = paste0("PC1 (", var_explained[1], "%)"),
  y = paste0("PC2 (", var_explained[2], "%)")) +
theme_classic(base_size = 14) +
theme(legend.position = "top")

pca_race
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#create a list of pluripotency, differentiation, and housekeeping genes
pluripotency_genes <- c("SOX2", "POU5F1", "KLF4", 
                        "MYC","PODXL", "NANOG", 
                        "FUT4", "LIN28A", "ALPL")

differentiation_genes <- c("PAX6","NES","KRT14", 
                           "SOX17", "AFP", "PDX1", 
                           "CD34", "THY1", "NKX2-5")

housekeeping_genes <- c("ACTB","GAPDH","RPL13A","HPRT1",
                        "B2M","TBP","PGK1","UBC", "RPL36AL")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#utilize a function to generate heatmap, lmm, and bar graph with ANOVA and Tukey HSD

run_gene_analysis <- function(gene_list, gene_set_name, condition = "age_sex") {
  
  # Filter logCPM to the selected genes and pivot long
  df <- logCPM_named %>%
    dplyr::select(-`Gene stable ID`) %>%
    filter(`Gene name` %in% gene_list) %>%
    pivot_longer(-`Gene name`, names_to = "Sample", 
                 values_to = "Expression") %>%
    left_join(
      raw_counts_cleaned %>% 
        dplyr::select(sample_id, sex, age_decade, age_sex, race),
      by = c("Sample" = "sample_id")) %>%
    filter(!is.na(age_decade) & !is.na(sex) & !is.na(race))
  
  df$group_var <- df[[condition]]

  #per gene ANOVA and Games_Howell
  
  # create empty dataframes to store results
  gene_anova <- data.frame()
  gene_Games_Howell <- data.frame()
  
#use a for loop to loop through each gene and perform an ANOVa and Games_Howell tests
  for (g in unique(df$`Gene name`)) {
    gdat <- df %>% filter(`Gene name` == g) %>%
    filter(!is.na(Expression), !is.na(age_sex), !is.na(race))
    
    if(nrow(gdat) < 3) next
    
    # One-way ANOVA (Expression ~ sex_age)
    fit <- oneway.test(Expression ~ group_var, 
                       data = gdat, var.equal = FALSE)
    
    # Save ANOVA result
    tidy_fit <- data.frame(
      term = condition,
      statistic = fit$statistic,
      parameter = fit$parameter,
      p.value = fit$p.value,
      Gene = g)
    gene_anova <- rbind(gene_anova, tidy_fit)
    
    # Games–Howell post-hoc
    gh <- gdat %>%
      rstatix::games_howell_test(Expression ~ group_var) %>%
      dplyr::rename(p_value = p.adj) %>%
      mutate(Gene = g,
             comparison = paste(group1, "-", group2))
    
    gene_Games_Howell <- rbind(gene_Games_Howell, gh)
  }

  # FDR correction for ANOVA
  anova_fdr <- gene_anova %>%
    group_by(term) %>%
    mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
    ungroup()

  # FDR correction for Games_Howell
  Games_Howell_fdr <- gene_Games_Howell %>%
    group_by(Gene) %>%
    mutate(p_adj = p.adjust(p_value, method = "BH"))
  
  # Summaries for plotting
  summary_df <- df %>%
    group_by(`Gene name`, group_var) %>%
    summarize(
      mean_expr = mean(Expression),
      SE_expr   = sd(Expression)/sqrt(n()),
      .groups = "drop") %>%
    rename(Gene = `Gene name`, `Sample name` = "group_var")
  
  summary_df <- summary_df %>%
    mutate(`Sample name` = trimws(`Sample name`),
         `Sample name` = factor(`Sample name`, 
                                levels = unique(`Sample name`)))
  
  # Significant pairs
  sig_pairs <- gene_Games_Howell%>%
    filter(p_value < 0.05) %>%
    mutate(
      annotations = case_when(
        p_value < 0.001 ~ "***",
        p_value < 0.01  ~ "**",
        TRUE ~ "*"),
      group1 = sub("-.*", "", comparison),
      group2 = sub(".*-", "", comparison)) %>%
    dplyr::select(Gene, group1, group2, p_value, annotations)
  
  sig_pairs <- sig_pairs %>%
  mutate(group1 = trimws(group1),
         group2 = trimws(group2),
         group1 = factor(group1, levels = levels(summary_df$`Sample name`)),
         group2 = factor(group2, levels = levels(summary_df$`Sample name`)))
  
  if (nrow(sig_pairs) > 0) {
    sig_pairs <- sig_pairs %>%
      left_join(
        summary_df %>%
          group_by(Gene) %>%
          summarize(
          max_expr = max(mean_expr + SE_expr),
          min_expr = min(mean_expr - SE_expr)),
      by = "Gene") %>%
    group_by(Gene) %>%
    mutate(
      y_position = ifelse(max_expr > 0,
                          max_expr * (1 + 1.1 * row_number()),  
                          min_expr * (1 - 1.1 * row_number()))) %>%
    ungroup()
  }
  
  # Bar graph
  plot <- ggplot(summary_df, 
                 aes(x = `Sample name`, 
                     y = mean_expr, 
                     fill = `Sample name`)) +
    geom_col(position =position_dodge(width =0.9)) +
    geom_errorbar(aes(ymin = mean_expr - SE_expr, 
                      ymax = mean_expr + SE_expr),
                      position = position_dodge(width =0.9), width =0.2) +
    facet_wrap(~ Gene, scales = "free_y") +
    scale_fill_d3("category20c") +
    theme_bw() +
    scale_y_continuous(expand =expansion(mult =c(0.02,0.2))) +
    theme(axis.text.x = element_text(angle =45,hjust =1),
          legend.position ="top",
          panel.spacing =unit(1.2,"lines"),
          panel.clip ="off",
          axis.text.y =element_text(size =12),
          axis.title.x =element_text(size =16),
          axis.title.y =element_text(size =16),
          strip.text =element_text(size =12,face ="bold")) +
    labs(y ="Log2(Counts Per Million)", 
         x ="Sample",
         fill ="Sample:")
  
  if (nrow(sig_pairs) > 0) {
    plot <- plot +
      geom_signif(
        data = sig_pairs,
        aes(xmin = group1, xmax = group2, y_position = y_position,
            annotations = annotations),
        manual = TRUE,
        inherit.aes = FALSE)
  }

  heatmap_mat <- df %>%
  group_by(`Gene name`, group_var) %>%
  summarize(Expression = mean(Expression, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = group_var, values_from = Expression) %>%
  column_to_rownames("Gene name") %>%
  as.matrix()

heatmap_obj <- pheatmap::pheatmap(
  heatmap_mat,
  scale = "row",                 
  cluster_rows = TRUE,           
  cluster_cols = FALSE,          
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 10,
  fontsize_col = 12,
  cellwidth = 50,                
  cellheight = 25,
  silent = TRUE)

  # Return everything
  list(
  data   = df,
  anova       = gene_anova,
  anova_fdr   = anova_fdr,
  Games_Howell = gene_Games_Howell,
  Games_Howell_fdr = Games_Howell_fdr,
  plot        = plot,
  heatmap = heatmap_obj)
}
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#run_gene_analysis function for each group of genes and contition

top20_agesex_results  <- run_gene_analysis(top_genes, "Top20", condition = "age_sex")
top20_race_results  <- run_gene_analysis(top_genes, "Top20", condition = "race")

pluri_agesex_results  <- run_gene_analysis(pluripotency_genes, "Pluripotency", condition = "age_sex")
pluri_race_results  <- run_gene_analysis(pluripotency_genes, "Pluripotency", condition = "race")

diff_agesex_results   <- run_gene_analysis(differentiation_genes, "Differentiation", condition = "age_sex")
diff_race_results   <- run_gene_analysis(differentiation_genes, "Differentiation", condition = "race")

housekeeping_agesex_results  <- run_gene_analysis(housekeeping_genes, "Housekeeping", condition = "age_sex")
housekeeping_race_results  <- run_gene_analysis(housekeeping_genes, "Housekeeping", condition = "race")
```

```{r fig-heatmap-top20, fig.cap = "Expression of Genes with the Greatest Variability", fig.width=8, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
top20_agesex_results$heatmap
```

```{r fig-heatmap-top20-race, fig.cap = "Expression of of Genes with the Greatest Variability", fig.width=8, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
top20_race_results$heatmap
```

```{r fig-heatmap-pluripotency, fig.cap = "Expression of Pluripotency Genes", fig.width=8, fig.height=5, echo=FALSE, warning=FALSE, message=FALSE}
pluri_agesex_results$heatmap
```

```{r fig-heatmap-pluripotency-race, fig.cap = "Expression of Pluripotency Genes", fig.width=8, fig.height=5, echo=FALSE, warning=FALSE, message=FALSE}
pluri_race_results$heatmap
```

```{r fig-bargraph-pluripotency, fig.cap = "Expression of Pluripotency Genes", fig.width=8, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
pluri_agesex_results$plot
```

```{r fig-bargraph-pluripotency-race, fig.cap = "Expression of Pluripotency Genes", fig.width=8, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
pluri_race_results$plot
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
#pluripotency genes, age sex stats 

df_pluri_agesex <- pluri_agesex_results$data
 # Fit mixed model
#model the response variable, Expression, for the fixed effects, age and sex
# genes are treated as a random effect, all with different y intercepts,
# since expression levels vary by gene
lmm_pluri_agesex <- lmer(Expression ~ age_sex + (1 | `Gene name`), data = df_pluri_agesex)
summary_pluri_agesex <- summary(lmm_pluri_agesex)
  # type III ANOVA to test significance of fixed effects on expression
lmm_pluri_agesex_anova <- anova(lmm_pluri_agesex, type = 3)

# estimated marginal means (or least-squares means) with pairwise comparisons
lmm_pluri_agesex_emm <- emmeans(lmm_pluri_agesex, pairwise ~ age_sex)

#lmm summary
summary_pluri_agesex
#anova
lmm_pluri_agesex_anova
#means 
lmm_pluri_agesex_emm
#gene anova FDR corrected 
pluri_agesex_results$anova_fdr
#gene tueky corrected
pluri_agesex_results$Games_Howell_fdr
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
#pluripotency genes, race stats 

df_pluri_race <- pluri_race_results$data
 # Fit mixed model
#model the response variable, Expression, for the fixed effects, age and sex
# genes are treated as a random effect, all with different y intercepts,
# since expression levels vary by gene
lmm_pluri_race <- lmer(Expression ~ race + (1 | `Gene name`), data = df_pluri_race)
summary_pluri_race <- summary(lmm_pluri_race)
  # type III ANOVA to test significance of fixed effects on expression
lmm_pluri_race_anova <- anova(lmm_pluri_race, type = 3)

# estimated marginal means (or least-squares means) with pairwise comparisons
lmm_pluri_race_emm <- emmeans(lmm_pluri_race, pairwise ~ race)

#lmm summary
summary_pluri_race
#anova
lmm_pluri_race_anova
#means 
lmm_pluri_race_emm
#gene anova FDR corrected 
pluri_race_results$anova_fdr
#gene tueky corrected
pluri_race_results$Games_Howell_fdr
```

```{r fig-heatmap-diff, fig.cap = "Expression of Differentiation Genes", fig.width=8, fig.height=5, echo=FALSE, warning=FALSE, message=FALSE}
diff_agesex_results$heatmap
```

```{r fig-heatmap-diff-race, fig.cap = "Expression of Differentiation Genes", fig.width=8, fig.height=5, echo=FALSE, warning=FALSE, message=FALSE}
diff_race_results$heatmap
```

```{r fig-bargraph-diff, fig.cap = "Expression of Differentiation Genes", fig.width=8, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
diff_agesex_results$plot
```

```{r fig-bargraph-diff-race, fig.cap = "Expression of Differentiation Genes", fig.width=8, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
diff_race_results$plot
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
#diff genes, age sex stats 

df_diff_agesex <- diff_agesex_results$data
 # Fit mixed model
#model the response variable, Expression, for the fixed effects, age and sex
# genes are treated as a random effect, all with different y intercepts,
# since expression levels vary by gene
lmm_diff_agesex <- lmer(Expression ~ age_sex + (1 | `Gene name`), data = df_diff_agesex)
summary_diff_agesex <- summary(lmm_diff_agesex)
  # type III ANOVA to test significance of fixed effects on expression
lmm_diff_agesex_anova <- anova(lmm_diff_agesex, type = 3)

# estimated marginal means (or least-squares means) with pairwise comparisons
lmm_diff_agesex_emm <- emmeans(lmm_diff_agesex, pairwise ~ age_sex)

#lmm summary
summary_diff_agesex
#anova
lmm_diff_agesex_anova
#means 
lmm_diff_agesex_emm
#gene anova FDR corrected 
diff_agesex_results$anova_fdr
#gene tueky corrected
diff_agesex_results$Games_Howell_fdr
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
#diff genes, race stats 

df_diff_race <- diff_race_results$data
 # Fit mixed model
#model the response variable, Expression, for the fixed effects, age and sex
# genes are treated as a random effect, all with different y intercepts,
# since expression levels vary by gene
lmm_diff_race <- lmer(Expression ~ race + (1 | `Gene name`), data = df_diff_race)
summary_diff_race <- summary(lmm_diff_race)
  # type III ANOVA to test significance of fixed effects on expression
lmm_diff_race_anova <- anova(lmm_diff_race, type = 3)

# estimated marginal means (or least-squares means) with pairwise comparisons
lmm_diff_race_emm <- emmeans(lmm_diff_race, pairwise ~ race)

#lmm summary
summary_diff_race
#anova
lmm_diff_race_anova
#means 
lmm_diff_race_emm
```

```{r fig-heatmap-hk, fig.cap = "Expression of Housekeeping Genes", fig.width=8, fig.height=5, echo=FALSE, warning=FALSE, message=FALSE}
housekeeping_agesex_results$heatmap
```

```{r fig-heatmap-hk-race, fig.cap = "Expression of Housekeeping Genes", fig.width=8, fig.height=5, echo=FALSE, warning=FALSE, message=FALSE}
housekeeping_race_results$heatmap
```

```{r fig-bargraph-hk, fig.cap = "Expression of Housekeeping Genes", fig.width=8, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
housekeeping_agesex_results$plot
```

```{r fig-bargraph-hk-race, fig.cap = "Expression of Housekeeping Genes", fig.width=8, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
housekeeping_race_results$plot
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
#hk genes, age sex stats 

df_housekeeping_agesex <- housekeeping_agesex_results$data
 # Fit mixed model
#model the response variable, Expression, for the fixed effects, age and sex
# genes are treated as a random effect, all with different y intercepts,
# since expression levels vary by gene
lmm_housekeeping_agesex <- lmer(Expression ~ age_sex + (1 | `Gene name`), data = df_housekeeping_agesex)
summary_housekeeping_agesex <- summary(lmm_housekeeping_agesex)
  # type III ANOVA to test significance of fixed effects on expression
lmm_housekeeping_agesex_anova <- anova(lmm_housekeeping_agesex, type = 3)

# estimated marginal means (or least-squares means) with pairwise comparisons
lmm_housekeeping_agesex_emm <- emmeans(lmm_housekeeping_agesex, pairwise ~ age_sex)

#lmm summary
summary_housekeeping_agesex
#anova
lmm_housekeeping_agesex_anova
#means 
lmm_housekeeping_agesex_emm
#gene anova FDR corrected 
housekeeping_agesex_results$anova_fdr
#gene tueky corrected
housekeeping_agesex_results$Games_Howell_fdr
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
#diff genes, race stats 

df_housekeeping_race <- housekeeping_race_results$data
 # Fit mixed model
#model the response variable, Expression, for the fixed effects, age and sex
# genes are treated as a random effect, all with different y intercepts,
# since expression levels vary by gene
lmm_housekeeping_race <- lmer(Expression ~ race + (1 | `Gene name`), data = df_housekeeping_race)
summary_housekeeping_race <- summary(lmm_housekeeping_race)
  # type III ANOVA to test significance of fixed effects on expression
lmm_housekeeping_race_anova <- anova(lmm_housekeeping_race, type = 3)

# estimated marginal means (or least-squares means) with pairwise comparisons
lmm_housekeeping_race_emm <- emmeans(lmm_housekeeping_race, pairwise ~ race)

#lmm summary
summary_housekeeping_race
#anova
lmm_housekeeping_race_anova
#means 
lmm_housekeeping_race_emm
#gene anova FDR corrected 
housekeeping_race_results$anova_fdr
#gene tueky corrected
housekeeping_race_results$Games_Howell_fdr
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
plot_lmm_residuals <- function(lmm, data, fixed_effect, extreme_sd = 2) {
  
  # Extract residuals
  residuals_df <- data %>%
    mutate(
      Residual = resid(lmm, type = "pearson"),
      Residual_SD = Residual / sd(Residual),
      Extreme = abs(Residual_SD) > extreme_sd)
  
  # Histogram of residuals
  p_hist <- ggplot(residuals_df, aes(x = Residual, fill = Extreme)) +
    geom_histogram(binwidth = 0.2, color = "black") +
    scale_fill_manual(values = c("grey80", "red")) +
    theme_bw() +
    labs(title = paste("Residuals from LMM:", fixed_effect),
         x = "Residual", y = "Count")
  
  # QQ plot of residuals
  p_qq <- ggplot(residuals_df, aes(sample = Residual)) +
    stat_qq(aes(color = Extreme)) +
    stat_qq_line() +
    scale_color_manual(values = c("black", "red")) +
    theme_bw() +
    labs(title = paste("QQ plot of LMM residuals:", fixed_effect),
         color = "Extreme Residual")
  
  # List extreme residuals
  extreme_residuals <- residuals_df %>%
    filter(Extreme) %>%
    select(`Gene name`, Sample, Residual) %>%
    arrange(desc(abs(Residual)))
  
  # Automatically display plots
  print(p_hist)
  print(p_qq)
    # Display extreme residuals dataframe
  print(extreme_residuals)
  
  #return data invisibly
  invisible(list(
    residuals_df = residuals_df,
    extreme_residuals = extreme_residuals))
}

#pluripotency genes
residuals_pluri_age_sex <- plot_lmm_residuals(
  lmm = lmm_pluri_agesex,
  data = df_pluri_agesex,
  fixed_effect = "age_sex",
  extreme_sd = 2)

residuals_pluri_race <- plot_lmm_residuals(
  lmm = lmm_pluri_race,
  data = df_pluri_race,
  fixed_effect = "race",
  extreme_sd = 2)

#differentiation genes
residuals_diff_age_sex <- plot_lmm_residuals(
  lmm = lmm_diff_agesex,
  data = df_diff_agesex,
  fixed_effect = "age_sex",
  extreme_sd = 2)

residuals_diff_race <- plot_lmm_residuals(
  lmm = lmm_diff_race,
  data = df_diff_race,
  fixed_effect = "race",
  extreme_sd = 2)

#housekeeping genes
residuals_housekeeping_age_sex <- plot_lmm_residuals(
  lmm = lmm_housekeeping_agesex,
  data = df_housekeeping_agesex,
  fixed_effect = "age_sex",
  extreme_sd = 2)

residuals_housekeeping_race <- plot_lmm_residuals(
  lmm = lmm_housekeeping_race,
  data = df_housekeeping_race,
  fixed_effect = "race",
  extreme_sd = 2)
```

# Discussion

### pie charts

The RNA Seq data is composed of a variety of different types of RNA, as
seen in Figure 1. 34.8% of read genes are mRNA protein coding genes,
whereas the remaining 65.2% of the detected genes are not protein coding
and are primary composed of noncoding RNAs and pseudogenes, which are
RNA sequences that look like protein coding genes but are not able to be
translated into a protein. 25.6% of the detected genes were
pseuodegenes. 25.5% were long non coding RNA sequences and 8.5% were
small RNAs, which included snoRNA, snRNA, scaRNA, sRNA, and ribozyme
RNA. 5.7% were "Other" RNAs, which include ribosoal RNA, immunoglobin
genes, T-cell receptor genes, and mitocondrial RNAs. The right pie
chart, on the other hand, displays the expression of these groups. For
each gene for each sample, a count per million number was provided by
Carcamo-Orive *et al.*. This is typically calculated by dividing the raw
count of the gene by the total reads in the sample and multiplying by
one million. The total expression was summed for each category, and this
was graphed in Figure 2, which displays that 91.7% of the expression is
from protein coding genes. The remaining 8.3% accounts for the
expression of small RNAs, long non coding RNA, pseuodgenes, and "Other"
genes. This is supported by \_\_\_ data from \_\_\_ article, which saw
\_\_\_\_.

### pca

welch's anova (one way analysis of means (not assuming equal variance)

PC1 and sex (p = 0.3892) PCA2 ( p = 0.6077)

PCA1 and age_Decade (p = 0.09443) and PCA2 (p = 0.2645)

PCA1 and age_sex (p = 0.1993) and PCA2 (p \< 0.001)

PCA1 and race (p \< 0.001) and PCA2 (p \< 0.001)

### top 20 genes:

sex linked RPS4Y1 Ribosomal protein S4, Y-linked. Present only in males.
High variance indicates sex is a strong driver. XIST X-inactive specific
transcript. Expressed in females for X-chromosome inactivation.
Sex-linked. DDX3Y DEAD-box helicase, Y-linked. Male-specific. TSIX
Antisense of XIST, regulates X-chromosome inactivation. Female-specific.
TXLNGY Y-linked, involved in transcription/translation regulation. KDM5D
Lysine demethylase 5D, Y-linked. Male-specific. USP9Y Ubiquitin-specific
peptidase 9, Y-linked. Male-specific. EIF1AY Eukaryotic translation
initiation factor 1A, Y-linked. Male-specific. UTY Ubiquitously
transcribed tetratricopeptide repeat, Y-linked. Male-specific. NLGN4Y
Neuroligin 4, Y-linked. Male-specific. RPS4XP22 Pseudogene similar to
RPS4X/RPS4Y.

other genes CST1 Cystatin SN, extracellular protease inhibitor. May vary
in differentiation. CER1 Cerberus, secreted factor involved in early
development and differentiation. LARGE1 Glycosyltransferase, involved in
post-translational modification. NLRP2 NLR family pyrin domain
containing 2, involved in early development, oocyte/embryo regulation.
COL3A1 Collagen type III alpha 1, ECM protein, differentiation-related.
ZNF208 Zinc finger protein, transcription factor, potential regulatory
role. APOA1 Apolipoprotein A1, lipid transport, may reflect metabolic
differences. MEG3 Long non-coding RNA, tumor suppressor, epigenetic
regulation, differentiation-related. LINC00261 Long non-coding RNA,
involved in endoderm differentiation.

### pluripotency

Expression of Pluripotency Genes (SOX2, POU5F1(OCT4), KLF4, MYC (c-Myc),
TRA-1-81 and 1-60 (PODXL), NANOG, FUT4 (SSEA4), LIN28A,alp

linear mixed model Expression \~ age_sex + 1\|`Gene name`

“We fit linear mixed-effects models of the form\

**Expression \~ age_sex + (1 \| Gene)**\

using lme4/lmerTest, with Satterthwaite-adjusted p-values.”

data: pluripotency genes

each gene has its own intercept, age_sex is a fixed effect

Residual variance = 0.189 and gene intercept variance = 9.82 means most
of the variation is between genes, not sex and age. the genes differ a
lot. residual noise is small, model fits wells

Fixed effects

70s female, lower pluripotency expression, - 0.097, p \< 0.05. 6.5%
decrease

70s male, higher pluripotency expression, +0.179, p \< 0.001 13%
increase

no difference between other groups and the reference group (40s female)

Type III Analysis of Variance Table with Satterthwaite's method -
confirms age_sex have a significant effect on pluripotency gene
expression when averaged across all genes. p \< 0.001.

FUT4 showed 11 significant pairwise comparisons across age-sex groups
after FDR correction (Games-Howell test), LIN28A showed 9, and NANOG,
POU5F1, and SOX2 with 5 signficant pairwiese comparisons.

KLF4 and ALPL showed the fewest significant differences, with only one
pairwise comparison reaching significance.

all gene has significant pairwise comparisons

linear mixed model Expression \~ race + 1\|`Gene name`

Residual variance = 0.1901 and gene intercept variance = 9.82 means most
of the variation is between genes, not race. the genes differ a lot.
residual noise is small, model fits wells

South Asian 2\^(-0.143) = 0.9 = 10% decrease in expression. (p \< 0.01)

After adjusting for gene-level random effects, South Asian individuals
show significantly lower mean log₂ gene expression (β = –0.143, p =
0.0059).\

This corresponds to approximately a 10% decrease in pluripotency gene
expression relative to the reference race group.

ALPL and LIN28A showed 5 significant pairwise comparisons, and SOX2 with
four.

Several pluripotency genes, including FUT4, NANOG, PODXL, and POU5F1,
did not show any significant pairwise differences across age-sex groups,
indicating relatively stable expression.

### Differentiation Markers

Expression of Differentiation Genes (ectoderm: PAX6, NESTIN (NES)
(neural projenitor/stem cell markers), KRT14 (epithelial), endoderm:
SOX17, AFP (hepatic), PDX1 (pancreatic), mesoderm: CD34 (hematopoitic),
CD90 (THY1)(mesenchymal), NKX2.5 (cardiac progenitor)).

age_sex

random effects:

Gene-level intercept variance = 13.168 (SD = 3.629), showing substantial
variation between genes.

Residual variance = 1.195 (SD = 1.093), within-gene variation.

fixed effects

50s female, -0.269, p \< 0.001 17% decrease

50s male -0.398, p\< 0.001 24.5% decrease

60s female -0.500, p \< 0.001 29.3% decrease

Compared with 40s female samples, differentiation gene expression was
17% lower in 50s female, 24.5% lower in 50s male, and 29.3% lower in 60s
female samples (p \< 0.001 for all). No significant differences were
observed for 40s male, 60s male, 70s female, or 70s male.”

race

Intercept 13.169 Residual1.163

Asian, -0.42975, 25.3% decrease p \< 0.001

Hispanic, + 0.39902, 32.0% increase, p \< 0.01

South Asian, + 0.85435, 80.0% increase, p \< 0.001

### Housekeeping Genes

Figure 5. Expression of Housekeeping Genes (some sort of general figure
or sorted by housekeeping type (i.e. ribosomal, metabolic, cytoskeleton,
etc.))

age_sex

Intercept 7.163 Residual 0.150

40s male, + 0.1325, 9.6% increase, p \< 0.001

50s female, + 0.03375, 2.4% increase, p \< 0.01

60s female, + 0.08191, 5.7% increase, p \< 0.01

race

Intercept 7.1635 Residual 0.1507

Hispanic, + 0.09717, 7.1% increase, p \< 0.05

# Conclusion

------------------------------------------------------------------------

# References
