---
title: "Gene Expression Variability in Human Induced Pluripotent Stem Cells"
subtitle: "BIOL 806 Final Project"
author: "Trinity Minard"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
format:
  pdf:
    keep-tex: true
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: inline
bibliography:
  - references.bib
  - packages.bib
csl: norsk-apa-manual.csl
---

# Abstract

Human induced pluripotent stem cells (iPSCs) are widely used to model
human development, disease, and for potential cell therapies. Accurate
characterization of iPSC lines is essential to ensure experimental and
manufacturing consistency, including assessing pluripotency,
differentiation, and reference gene expression. However, previous
studies have shown that commonly used housekeeping genes, such as GAPDH,
can vary across iPSC lines, complicating normalization.

Here, we analyzed publicly available RNA-seq data from 317 iPSC samples
from 101 diverse donors to quantify RNA composition, identify drivers of
gene expression variability, and evaluate pluripotency, differentiation,
and housekeeping gene expression. While multiple RNA types were
detected, 91.7% of total expression originated from protein-coding
mRNAs. Principal component analysis (PCA) and Welch’s one-way ANOVA
revealed that donor age–sex groups and race contributed significantly to
expression variability, with sex-linked genes among the top drivers.
Pluripotency markers, such as LIN28A and SOX2, and differentiation
markers, such as SOX17 and CD34, amongst many other genes, exhibited
variable expression across donor groups. Notably, housekeeping genes
(GAPDH, HPRT1, PGK1) also significantly varied, highlighting limitations
for their use as universal references.

These results highlight that donor age, sex, and race can significantly
influence iPSC transcriptomes, affecting the expression of many genes
that are commonly used as references and control markers. This displays
the need to account for donor demographics when interpreting gene
expression data, selecting reference genes, or designing experiments
using iPSCs, to ensure accurate and reproducible results.

# Introduction

The discovery of human induced pluripotent stem cells (iPSCs) by
Takahashi and Yamanaka in 2007 [@takahashi2007] has revolutionized the
regenerative medicine field. iPSCs are created by reprogramming somatic
cells into an embryonic like state using the Yamanaka transciption
factors: SOX2, OCT4, KLF4, and c-Myc. These cells have vast capabilities
due to their ability to model human development and diseases *in vitro*,
and their potential for patient specific (autologous) and "off the
shelf" (allogeneic) cell therapies [@cerneckis2024].

With that being said, the characterization of iPSC lines is essential in
any experimental or manufacturing process to ensure consistency in
culturing. This includes analyzing the expression of pluripotency (stem
cell) markers, differentiation markers, and potential chromosomal
abnormalies, to access the population of the culture. This is commonly
done using techniques such as reverse transcriptase or real time
quantitative PCR (RT-qPCR) to analyze mRNA expression [@artyukhov2017],
or using an antibody based experiment to analyze protein expression
levels, such as Flow Cytometry or immunocytochemistry. When conducting
RT-qPCR and similar experiments, reference genes, also known as
housekeeping genes, are used to normalize the expression of other genes
in a sample [@artyukhov2017]. This allows the expression of genes to be
compared across different samples, since it is presumed that the
expression of the housekeeping gene is consistent across all of the
samples. In stem cell samples, however, it has been found that the
commonly used housekeeping genes, such as GAPDH, appear to fluctuate in
iPSC samples [@panina2020], making it difficult to achieve accurate
normalization data.

RNA-Seq is a technique that sequences the entirety of RNA molecules in a
given sample, rather than using gene specific primers or probes as in
RT-qPCR and other techniques. This allows the entire transciptome to be
analyzed, providing a complete picture of the expression of genes in a
sample [@wang2009]. Additionally, gene expression is calculated by
dividing the raw counts of a gene by the total counts in a sample,
eliminating the need for a housekeeping gene. This allows for an
unbiased and quantitiative analysis of millions of genes in sample
[@wang2009]. Despite these advantages, RNA-Seq is far more time
consuming and expensive than RT-qPCR and other experiments, making it
less practical in a manufacturing setting. Therefore, utilizing RNA-Seq
to analyze a large number of iPSC samples could provide insight into
housekeeping gene expression, allowing researchers to select ideal
reference genes for their analysis.

A similar study, conducted by Carcamo-Orive *et al.,* investigated the
variability of gene expression across 317 iPSC lines generated from 101
individuals [@carcamo-orive2017] and created a publically avaible data
set of their raw RNA Seq data. They found that genetic variability in
the iPSC lines was influenced by a donor specific gene expression
patterns, specifically influenced by donor age, sex, ancestry, and
reprogramming variabilies. iPSCs can also be reprogrammed from a variety
of somatic cells, such as blood, skin fibroblasts, hair keratinocytes,
urine cells, mesencymal stem cells, and bone marrow [@poetsch2022].
While some somatic cells are easier to obtain than others, such as blood
cells and fibroblasts, iPSCs from these lines are associated with higher
genomic mutations due to the somatic cells' high turnover rate.
Additionally, iPSCs appear to have an epigenetic memory of their tissue
of origin, which has been found to influence their differentiation
[@poetsch2022]. All of these factors, and many others, could be
contributing to the variations in gene expression in iPSC samples.

This publicly available dataset from Carcamo-Orive *et al.* was used to
analyze the expression of RNA across iPSC samples, with a focus on the
expression of pluripotency markers, differentiation markers, and
housekeeping genes.

# Methods

```{r packages_references, echo=FALSE, warning=FALSE, message=FALSE}
#create list of packages to cite
packages <- c("base", "tidyverse", "dplyr", "ggplot2", "scales", "pheatmap", "lme4", "lmerTest", "emmeans", "broom", "ggsignif", "ggsci", "patchwork", "FactoMineR", "grid", "factoextra", "rstatix", "kableExtra", "lattice", "knitr", "gridExtra")

#citing R and packages
knitr::write_bib(packages, "packages.bib")
```

All scripts used for processing, filtering, visualization, and
statistical analysis are available at:
<https://github.com/tfm1017/BIOL806-Final-Project>.

### Data Cleaning

RNA-Seq counts per million data (named: cpm_data), and sample metadata
(named: raw_counts) were acquired from Stemformatics using the
Carcamo-Orive (2017) dataset [@carcamo-orive2017]. Genes in this study
were referred to by their Ensembl ID, so gene information, such as name
and RNA type, was integrated from the Ensembl database. Using R
[@R-base], the raw count per million data was converted to a log base 2
scale with a 0.5 offset to prevent samples with a count of 0 from
becoming skewed. Values for age, from the metadata file, were grouped by
decade, and an additional column was added for an age + sex category
(i.e., 40s female). Information from the sample description was then
extracted to create columns for race and insulin resistance. The
metadata and log₂ counts per million files were then merged into one
data frame for further analysis.

### Category Tables

The number of samples for each group within the age-sex and race
categories were summarized and displayed as two tables using the knitr
[@R-knitr] and kableExtra [@R-kableExtra] packages.

### Pie Charts by RNA Category

Genes detected in the RNA Seq counts per million data were grouped into
the following categories: Protein coding, long non-coding RNA (lncRNA),
Small RNAs, Pseudogenes, and Other. A count of the number of genes in
each category was then performed by using the dplyr package [@R-dplyr]
from tidyverse [@R-tidyverse]. The total expression of each category was
also calculated using the raw count per million data. The scales
[@R-scales] package from ggplot2 [@R-ggplot2] was used to create two pie
charts, one of the gene counts and one of the gene expression by RNA
category, plotted together using the patchwork package [@R-patchwork].

### Principal Component Analysis (PCA) Plots

Log2 expression values were first cleaned by removing entries without
gene information and duplicate gene names. Gene variances were
calculated across the samples, and genes with no variance were removed.
To visualize the major sources of variation in the expression data, a
principal component analysis (PCA) was performed using the FactoMineR
package [@R-FactoMineR], which identifies the major axes of variation in
the dataset. To assess if sample characteristics (sex, age decade,
combined age sex groups, and race) influenced the PCA’s structure, a
Welch’s one-way ANOVA, which does not assume equal variance, was
performed separately for PCA1 and PCA2. Significant differences were
found for the age sex groups and race categories, so two PCA biplots
were generated, each overlaying confidence ellipses corresponding to
either age sex groups or race. On both of the biplots, the top twenty
genes with the highest variance across all samples were labeled and
displayed as vectors. Additionally, PCA data was plotted as four
scatterplots, colored by age, sex, age-sex, and race.

### Pluripotency, Differentiation, and Housekeeping Gene Expression

The log2 expression of genes, categorized as pluripotency markers,
differentiation markers, and housekeeping genes, was compared across
sample groups (age-sex and race) using heatmaps, linear mixed-effects
models, and Welch’s ANOVA with post hoc Games-Howell tests. Heatmaps of
log2 gene expression were generated using the pheatmap package
[@R-pheatmap] to visualize overall expression patterns within each gene
category. Linear mixed-effect models were then fit using lme4 [@R-lme4]
and lmerTest [@R-lmerTest] to evaluate differences in gene expression
across age, sex, age-sex ,and race groups, while accounting for
gene-specific baseline expression levels by using random intercepts for
genes. Percent differences in overall gene expression were calculated by
exponentiating the estimated log2 scale coefficient back to fold change,
subtracting one, and multiplying by 100. QQ plots and histograms were
generated, but not shown here, for each model to view residuals and
assess model assumptions. To identify which genes displayed
statistically significant differences in expression between sample
categories, Welch’s one-way ANOVA, assuming unequal variance, was
performed for each gene. P values were adjusted using the
Benjamini-Hochberg method due to the high number of tests that were
performed. For genes with significant ANOVA results, Games-Howell post
hoc tests, using rstatix [@R-rstatix], were performed and p-values were
Benjamini-Hochberg corrected. Significant Games Howell comparisons
(adjusted p \< 0.05) were annotated on bar plots displaying log2
expression for each gene across sample groups.

```{r import_packages_and_data, echo=FALSE, message=FALSE, warning=FALSE}
#library packages
library(tidyverse)
library(dplyr)
library(scales)
library(ggrepel)
library(pheatmap)
library(lme4)
library(lmerTest)  
library(emmeans)
library(broom)
library(ggplot2)
library(ggsignif)
library(ggsci)
library(patchwork)
library(FactoMineR)
library(grid) 
library(gridExtra)
library(factoextra)
library(rstatix)
library(kableExtra)
library(lattice)
library(knitr)

#import data files
raw_counts <- read_tsv("stemformatics_dataset_7274.samples.tsv")
cpm_data   <- read_tsv("stemformatics_dataset_7274.cpm.tsv")
ensembl_data <- read_csv("mart_export.txt")

# view data files
#head(raw_counts)
#head(cpm_data)
```

```{r filter_genes, echo=FALSE, warning=FALSE, message=FALSE}
#Change column name from "0.001" to "Gene Stable ID"
cpm_data <- cpm_data %>%
  rename("Gene stable ID" = "...1")

#Create a data frame with only the gene info for the genes in this study
ensembl_filtered <- ensembl_data %>%
  filter(`Gene stable ID` %in% cpm_data$`Gene stable ID`)

```

```{r categorizing_RNA, echo=FALSE, warning=FALSE, message=FALSE}
# Add a new column 'Category' based on 'Gene type'
ensembl_filtered <- ensembl_filtered %>%
  mutate(Category = case_when(
    `Gene type` %in% c("protein_coding") ~ "Protein coding",
    `Gene type` %in% c("lncRNA", "TEC") ~ "lncRNA",
    `Gene type` %in% c("miRNA", 
                          "snoRNA", 
                          "snRNA", 
                          "scaRNA", 
                          "sRNA", 
                          "vault_RNA", 
                          "ribozyme") ~ "Small RNAs",
    `Gene type` %in% c("pseudogene", 
                          "processed_pseudogene", 
                          "unprocessed_pseudogene",
                          "transcribed_unprocessed_pseudogene",
                          "transcribed_processed_pseudogene",
                          "unitary_pseudogene", 
                          "IG_C_pseudogene", 
                          "IG_V_pseudogene",
                          "IG_J_pseudogene", 
                          "TR_J_pseudogene", 
                          "TR_V_pseudogene",
                          "translated_processed_pseudogene",
                        "transcribed_unitary_pseudogene") ~ "Pseudogenes",
       `Gene type` %in% c("misc_RNA", 
                          "artifact", 
                          "Mt_tRNA", 
                          "IG_V_gene", 
                          "IG_J_gene", 
                          "IG_C_gene", 
                          "IG_D_gene",
                          "TR_J_gene", 
                          "TR_V_gene", 
                          "TR_D_gene", 
                          "TR_C_gene",
                          "rRNA", 
                          "Mt_rRNA", 
                          "rRNA_pseudogene") ~ "Other",
   TRUE ~ "Other"))

```

```{r pie_chart_summarizing, echo=FALSE, warning=FALSE, message=FALSE}
#summarize count of genes by RNA category
RNA_types_summary <- ensembl_filtered %>%
  count(Category) %>%  # creates a numeric column `n`
  mutate(
    fraction   = n / sum(n),
    label_text = paste0(Category, "\n", round(fraction * 100, 1), "%"),
    large_slice = fraction > 0.1)

# pivot raw cpm data and add ensembl data information by Gene. then remove NAs
cpm_long <- cpm_data %>%
  pivot_longer(-`Gene stable ID`, names_to = "Sample", values_to = "Expression") %>%
  left_join(ensembl_filtered, by = "Gene stable ID") %>%
  filter(!is.na(Category), !is.na(Expression)) 

# summarize expression per category 
cpm_category_summary <- cpm_long %>%
  group_by(Category) %>%
  summarize(Total_Expression = sum(Expression)) %>%
  ungroup() %>%
  mutate(
    fraction = Total_Expression / sum(Total_Expression),
    label_text = paste0(Category, "\n", round(fraction * 100, 1), "%"),
    label_pos = cumsum(fraction) - fraction / 2,
    large_slice = fraction > 0.1) 
```

```{r ggplot_pie_charts, echo=FALSE, warning=FALSE, message=FALSE}

#manually set colors for each category (using factors was causing issues when plotting)
slice_colors <- c(
  "Protein coding" = "#7570b3",
  "Pseudogenes"    = "#d95f02",
  "lncRNA"         = "#1b9e77",
  "Small RNAs"     = "#e7298a",
  "Other"          = "#66a61e")

#pie chart of RNA categories 
plot_RNA_types <- ggplot(RNA_types_summary, aes(x = "", y = fraction, fill = Category)) +
  geom_bar(stat="identity", width=1, color="white") +
  scale_fill_manual(values = slice_colors) +
  coord_polar(theta="y") +
  theme_void() +
  theme(
    legend.position = "none",
    plot.margin = margin(0,0,0,0))

# Extract actual slice positions
pb <- ggplot_build(plot_RNA_types)
bar_positions <- pb$data[[1]]  # first layer = geom_bar
bar_positions$Category <- RNA_types_summary$Category
bar_positions$label_text <- RNA_types_summary$label_text
bar_positions$large_slice <- RNA_types_summary$large_slice

# Compute label positions for all slices
bar_positions <- bar_positions %>%
  mutate(label_pos = (ymin + ymax)/2)

# Add labels
plot_RNA_types_final <- plot_RNA_types +
  # On-slice labels for large slices
  geom_text(
    data = subset(bar_positions, large_slice),
    aes(y = label_pos, label = label_text),
    color = "white",
    fontface = "bold",
    size = 4) +
  # Repelled labels for small slices
  geom_text_repel(
    data = subset(bar_positions, !large_slice),
    aes(x = 1, y = label_pos, label = label_text),
    nudge_x = 0.8,
    force = 0.5,
    segment.color = "grey60",
    segment.length = 0.2,
    size = 4,
    fontface = "bold",
    inherit.aes = FALSE) +
  labs(caption = "Number of Genes per RNA Category") +
  theme(plot.caption = element_text(size=12, hjust=0.5, face="bold"))

#pie chart of RNA expression
plot_expression <- ggplot(cpm_category_summary, aes(x = "", y = fraction, fill = Category)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  scale_fill_manual(values = slice_colors) +
  coord_polar(theta = "y") +
  theme_void() +
  theme(
    legend.position = "none",
    plot.margin = margin(0, 0, 0, 0))

# Extract actual slice positions from ggplot
pb <- ggplot_build(plot_expression)
bar_positions <- pb$data[[1]]  # first layer = geom_bar
bar_positions$Category <- cpm_category_summary$Category
bar_positions$label_text <- cpm_category_summary$label_text
bar_positions$large_slice <- cpm_category_summary$large_slice

# Compute label positions
bar_positions <- bar_positions %>%
  mutate(label_pos = (ymin + ymax)/2)

# Add labels
plot_RNA_expression_final <- plot_expression +
  # On-slice labels for large slices
  geom_text(
    data = subset(bar_positions, large_slice),
    aes(y = label_pos, label = label_text),
    color = "white",
    fontface = "bold",
    size = 4) +
  # Repelled labels for small slices
  geom_text_repel(
    data = subset(bar_positions, !large_slice),
    aes(x = 1, y = label_pos, label = label_text),
    nudge_x = 0.8,
    force = 0.5,
    segment.color = "grey60",
    segment.length = 0.2,
    size = 4,
    fontface = "bold",
    inherit.aes = FALSE) +
  labs(caption = "Proportion of Total Expression per RNA Category") +
  theme(plot.caption = element_text(size = 12, hjust = 0.5, face = "bold"))
```

```{r data_cleaning, echo=FALSE, message=FALSE, warning=FALSE}
#begin cleaning data 

#verify that the sample ids match in raw_counts and _cpm_data
raw_counts <- raw_counts[match(colnames(cpm_data)[-1], raw_counts$sample_id), ]

#all(colnames(cpm_data)[-1] == raw_counts$sample_id)

#convert counts per million to a log scale, excluding the first column which has gene info, and adding an offset so that values use to 0 are not skewed
logCPM <- cbind(`Gene stable ID` = cpm_data$`Gene stable ID`, log2(cpm_data[,-1] + 0.5))

#summary(raw_counts)
#unique(raw_counts$parental_cell_type) #all parent cells are blood mononuclear
#unique(raw_counts$disease_state) #all normal
#unique(raw_counts$sex) #male and female
#unique(raw_counts$developmental_stage) #33 ages 

#based on information regarding the samples, sex, age and race differ across the samples.

#age is reported as "age years old", so a new column was added with just the numeric
raw_counts$age_numeric <- as.numeric(sub("-year-old.*", "", raw_counts$developmental_stage))

#age was then grouped by decade to reduce the number of conditions 
raw_counts$age_decade <- cut(
  raw_counts$age_numeric,
  breaks = seq(40, 80, by = 10),
  labels = c("40s", "50s", "60s", "70s"),
  right = FALSE)

#create a new factor column in the raw counts data which combined age and sex 
raw_counts_cleaned <- raw_counts %>%
 filter(!is.na(age_decade) & !is.na(sex)) %>%
  mutate(age_sex = factor(paste(age_decade, sex, sep = " "))) %>%
  mutate(
    # iPSC line ID (number_number)
    line_id = str_extract(sample_description, "\\d+_\\d+"),
    
    # sex: look for "female" or "male"
    sex = str_extract(sample_description, "female|male"),
    
    # age in years
    ageYr = as.numeric(str_extract(sample_description, "\\d+(?=yr)")),
    
    # race
    race = str_extract(sample_description, "caucasian|Asian|South Asian|African American|Hispanic"),
    
    # insulin status: "Insulin sensitive" or "Insulin resistant"
    insulin_status = str_extract(sample_description, "Insulin [a-zA-Z]+")) %>%
  mutate(
    sex = factor(sex),
    race = factor(race),
    insulin_status = factor(insulin_status)) %>%
  mutate(race = recode(race,
                       "caucasian" = "Caucasian"))

#select columns in raw counts which have minimal NAs and are variable
raw_counts_cleaned <- raw_counts_cleaned %>%
  select(sample_id, sex, reprogramming_method, 
         external_source_id, age_numeric, age_decade, 
         age_sex, race, insulin_status)

# add gene name from ensembl filtered to longCPM data
logCPM_named <- logCPM %>%
  left_join(
    ensembl_filtered %>% dplyr::select(`Gene stable ID`, `Gene name`),
    by = c("Gene stable ID" = "Gene stable ID"))

# move gene name to first column
logCPM_named <- logCPM_named %>%
  dplyr::select(`Gene name`, everything())
```

```{r pca, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
#remove rows from log_CPM_named when gene is NA and remove duplicates
logCPM_named_clean <- logCPM_named %>%
  filter(!is.na(`Gene name`) & `Gene name` != "") %>%
  distinct(`Gene name`, .keep_all = TRUE) 

#create a matrix of the expression data excluding gene info 
mat <- logCPM_named_clean %>%
  select(where(is.numeric)) %>%
  as.matrix()

#set each row in the matrix as its corresponding gene name 
rownames(mat) <- logCPM_named_clean$`Gene name`

#calculates gene variance 
gene_var <- apply(mat, 1, var)

#removes genes with minimal variance to reduce clutter in the plot
mat_filtered <- mat[gene_var > 0, ]

#selects the top 20 genes with the most variance to label in the plot
top_n <- 20  # number of genes to show as arrows 

top_genes <- names(sort(gene_var[gene_var > 0], decreasing = TRUE))[1:top_n]

mat_top <- mat_filtered[top_genes, ]

mat_t_top <- t(mat_top)

#turn top genes into a data frame
top_genes_df <- data.frame(
  Gene = top_genes,
  Variance = gene_var[top_genes]) %>%
  arrange(desc(Variance))

#transpose the matrix to have samples as rows and genes as columns for FactoMineR
mat_t <- t(mat_filtered)  

#perform principal component analysis 
pca_res <- PCA(mat_t, graph = FALSE)

#create sample coordinates and add them as columns PC1 and PC2
pca_df <- data.frame(pca_res$ind$coord[,1:2])

colnames(pca_df) <- c("PC1", "PC2")

pca_df$Sample <- rownames(pca_df)

#add sample info from raw_counts to the pca dataframe 
pca_df <- pca_df %>%
  left_join(raw_counts_cleaned, by = c("Sample" = "sample_id"))

#calculate percent variance 
var_explained <- round(pca_res$eig[1:2, "percentage of variance"], 1)
```

```{r pca-graph, echo=FALSE, warning=FALSE, message=FALSE}
# extract loadings
loadings <- as.data.frame(pca_res$var$coord)

# add gene names
loadings$Gene <- rownames(loadings)

# rank genes by contribution to PC1
top_PC1_genes <- loadings %>%
  arrange(desc(abs(Dim.1))) %>%
  slice(1:20) %>%
  select(Gene, Dim.1)
#print(top_PC1_genes)

# rank genes by contribution to PC2
top_PC2_genes <- loadings %>%
  arrange(desc(abs(Dim.2))) %>%
  slice(1:20) %>%
  select(Gene, Dim.2)
#print(top_PC2_genes)

PCA_sex <- ggplot(pca_df, aes(PC1, PC2, color = sex)) +
  geom_point(size = 3) +
  labs(
  x = paste0("PC1 (", var_explained[1], "%)"),
  y = paste0("PC2 (", var_explained[2], "%)"),
  color = "Group:") +
  theme_classic(base_size = 14) +
  theme(legend.position = "top")

PCA_age <- ggplot(pca_df, aes(PC1, PC2, color = age_decade)) +
  geom_point(size = 3) +
  labs(
  x = paste0("PC1 (", var_explained[1], "%)"),
  y = paste0("PC2 (", var_explained[2], "%)"),
  color = "Group:") +
  theme_classic(base_size = 14) +
  theme(legend.position = "top")

PCA_age_sex <- ggplot(pca_df, aes(PC1, PC2, color = age_sex)) +
  geom_point(size = 3) +
  labs(
  x = paste0("PC1 (", var_explained[1], "%)"),
  y = paste0("PC2 (", var_explained[2], "%)"),
  color = "Group:") +
  theme_classic(base_size = 14) +
  theme(legend.position = "top")

PCA_race <- ggplot(pca_df, aes(PC1, PC2, color = race)) +
  geom_point(size = 3) +
  labs(
  x = paste0("PC1 (", var_explained[1], "%)"),
  y = paste0("PC2 (", var_explained[2], "%)"),
  color = "Group:") +
  theme_classic(base_size = 14) +
  theme(legend.position = "top")
```

```{r pca_anovas, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
# use anova to see if sex, age, or race influence PC1 or PC2 

#sex alone
oneway.test(PC1 ~ sex, data = pca_df, var.equal = FALSE)
oneway.test(PC2 ~ sex, data = pca_df, var.equal = FALSE)

#age alone
oneway.test(PC1 ~ age_decade, data = pca_df, var.equal = FALSE)
oneway.test(PC2 ~ age_decade, data = pca_df, var.equal = FALSE)

#age and decade
oneway.test(PC1 ~ age_sex, data = pca_df, var.equal = FALSE)
oneway.test(PC2 ~ age_sex, data = pca_df, var.equal = FALSE)

#race alone
oneway.test(PC1 ~ race, data = pca_df, var.equal = FALSE)
oneway.test(PC2 ~ race, data = pca_df, var.equal = FALSE)
```

```{r gene_lists, echo=FALSE, message=FALSE, warning=FALSE}
#create a list of pluripotency, differentiation, and housekeeping genes
pluripotency_genes <- c("SOX2", "POU5F1", "KLF4", 
                        "MYC","PODXL", "NANOG", 
                        "FUT4", "LIN28A", "ALPL")

differentiation_genes <- c("PAX6","NES","KRT14", 
                           "SOX17", "AFP", "PDX1", 
                           "CD34", "THY1", "NKX2-5")

housekeeping_genes <- c("ACTB","GAPDH","RPL13A","HPRT1",
                        "B2M","TBP","PGK1","UBC", "RPL36AL")
```

```{r gene_analysis_function, echo=FALSE, message=FALSE, warning=FALSE}
#utilize a function to generate heatmap, lmm, and bar graph with ANOVA and Tukey HSD

run_gene_analysis <- function(gene_list, gene_set_name, condition = "age_sex") {
  
  # Filter logCPM to the selected genes and pivot long
  df <- logCPM_named %>%
    dplyr::select(-`Gene stable ID`) %>%
    filter(`Gene name` %in% gene_list) %>%
    pivot_longer(-`Gene name`, names_to = "Sample", 
                 values_to = "Expression") %>%
    left_join(
      raw_counts_cleaned %>% 
        dplyr::select(sample_id, sex, age_decade, age_sex, race),
      by = c("Sample" = "sample_id")) %>%
    filter(!is.na(age_decade) & !is.na(sex) & !is.na(race))
  
  df$group_var <- df[[condition]]

  #per gene ANOVA and Games_Howell
  
  # create empty dataframes to store results
  gene_anova <- data.frame()
  gene_Games_Howell <- data.frame()
  
#use a for loop to loop through each gene and perform an ANOVa and Games_Howell tests
  for (g in unique(df$`Gene name`)) {
    gdat <- df %>% filter(`Gene name` == g) %>%
    filter(!is.na(Expression), !is.na(age_sex), !is.na(race))
    
    if(nrow(gdat) < 3) next
    
    # One-way ANOVA (Expression ~ sex_age)
    fit <- oneway.test(Expression ~ group_var, 
                       data = gdat, var.equal = FALSE)
    
    # Save ANOVA result
    tidy_fit <- data.frame(
      term = condition,
      statistic = fit$statistic,
      parameter = fit$parameter,
      p.value = fit$p.value,
      Gene = g)
    gene_anova <- rbind(gene_anova, tidy_fit)
    
    # Games–Howell post-hoc
    gh <- gdat %>%
      rstatix::games_howell_test(Expression ~ group_var) %>%
      dplyr::rename(p_value = p.adj) %>%
      mutate(Gene = g,
             comparison = paste(group1, "-", group2))
    
    gene_Games_Howell <- rbind(gene_Games_Howell, gh)
  }

  # FDR correction for ANOVA
  anova_fdr <- gene_anova %>%
    group_by(term) %>%
    mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
    ungroup()

  # FDR correction for Games_Howell
  Games_Howell_fdr <- gene_Games_Howell %>%
    group_by(Gene) %>%
    mutate(p_adj = p.adjust(p_value, method = "BH"))
  
  # Summaries for plotting
  summary_df <- df %>%
    group_by(`Gene name`, group_var) %>%
    summarize(
      mean_expr = mean(Expression),
      SE_expr   = sd(Expression)/sqrt(n()),
      .groups = "drop") %>%
    rename(Gene = `Gene name`, `Sample name` = "group_var")
  
  summary_df <- summary_df %>%
    mutate(`Sample name` = trimws(`Sample name`),
         `Sample name` = factor(`Sample name`, 
                                levels = unique(`Sample name`)))
  
  # Significant pairs
  sig_pairs <- gene_Games_Howell%>%
    filter(p_value < 0.05) %>%
    mutate(
      annotations = case_when(
        p_value < 0.001 ~ "***",
        p_value < 0.01  ~ "**",
        TRUE ~ "*"),
      group1 = sub("-.*", "", comparison),
      group2 = sub(".*-", "", comparison)) %>%
    dplyr::select(Gene, group1, group2, p_value, annotations)
  
  sig_pairs <- sig_pairs %>%
  mutate(group1 = trimws(group1),
         group2 = trimws(group2),
         group1 = factor(group1, levels = levels(summary_df$`Sample name`)),
         group2 = factor(group2, levels = levels(summary_df$`Sample name`)))
  
  if (nrow(sig_pairs) > 0) {
    sig_pairs <- sig_pairs %>%
      left_join(
        summary_df %>%
          group_by(Gene) %>%
          summarize(
          max_expr = max(mean_expr + SE_expr),
          min_expr = min(mean_expr - SE_expr)),
      by = "Gene") %>%
    group_by(Gene) %>%
    mutate(
      y_position = ifelse(max_expr > 0,
                          max_expr * (1 + 1.1 * row_number()),  
                          min_expr * (1 - 1.1 * row_number()))) %>%
    ungroup()
  }
  
  # Bar graph
  plot <- ggplot(summary_df, 
                 aes(x = `Sample name`, 
                     y = mean_expr, 
                     fill = `Sample name`)) +
    geom_col(position =position_dodge(width =0.9)) +
    geom_errorbar(aes(ymin = mean_expr - SE_expr, 
                      ymax = mean_expr + SE_expr),
                      position = position_dodge(width =0.9), width =0.2) +
    facet_wrap(~ Gene, scales = "free_y") +
    scale_fill_d3("category20c") +
    theme_bw() +
    scale_y_continuous(expand =expansion(mult =c(0.02,0.2))) +
    theme(axis.text.x = element_text(angle =45,hjust =1),
          legend.position ="top",
          panel.spacing =unit(1.2,"lines"),
          panel.clip ="off",
          axis.text.y =element_text(size =12),
          axis.title.x =element_text(size =16),
          axis.title.y =element_text(size =16),
          strip.text =element_text(size =12,face ="bold")) +
    labs(y ="Expression (log2(CPM))",
         fill ="Sample:")
  
  if (nrow(sig_pairs) > 0) {
    plot <- plot +
      geom_signif(
        data = sig_pairs,
        aes(xmin = group1, xmax = group2, y_position = y_position,
            annotations = annotations),
        manual = TRUE,
        inherit.aes = FALSE)
  }

  heatmap_mat <- df %>%
  group_by(`Gene name`, group_var) %>%
  summarize(Expression = mean(Expression, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = group_var, values_from = Expression) %>%
  column_to_rownames("Gene name") %>%
  as.matrix()

heatmap_obj <- pheatmap::pheatmap(
  heatmap_mat,
  scale = "row",                 
  cluster_rows = TRUE,           
  cluster_cols = FALSE,          
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 10,
  fontsize_col = 12,
  cellwidth = 50,                
  cellheight = 25,
  silent = TRUE)

  # Return everything
  list(
  data   = df,
  anova       = gene_anova,
  anova_fdr   = anova_fdr,
  Games_Howell = gene_Games_Howell,
  Games_Howell_fdr = Games_Howell_fdr,
  plot        = plot,
  heatmap = heatmap_obj)
}
```

```{r run_gene_analysis, echo=FALSE, message=FALSE, warning=FALSE}
#run_gene_analysis function for each group of genes and contition

top20_agesex_results  <- run_gene_analysis(top_genes, "Top20", condition = "age_sex")
top20_race_results  <- run_gene_analysis(top_genes, "Top20", condition = "race")

pluri_agesex_results  <- run_gene_analysis(pluripotency_genes, "Pluripotency", condition = "age_sex")
pluri_race_results  <- run_gene_analysis(pluripotency_genes, "Pluripotency", condition = "race")

diff_agesex_results   <- run_gene_analysis(differentiation_genes, "Differentiation", condition = "age_sex")
diff_race_results   <- run_gene_analysis(differentiation_genes, "Differentiation", condition = "race")

housekeeping_agesex_results  <- run_gene_analysis(housekeeping_genes, "Housekeeping", condition = "age_sex")
housekeeping_race_results  <- run_gene_analysis(housekeeping_genes, "Housekeeping", condition = "race")
```

# Results

```{r count_tables, echo=FALSE, warning=FALSE, message=FALSE}
#view n numbers by category and create kable tables 
raw_counts_cleaned %>% 
  count(age_sex) %>% 
  arrange(age_sex) %>% 
  rename(`Age-Sex Group` = age_sex, `Sample Count` = n) %>% 
  kable("latex", booktabs = TRUE, caption = "Sample counts by Age-Sex") %>% 
  kable_styling(latex_options = c("striped", "hold_position"))

raw_counts_cleaned %>% 
  count(race) %>% 
  arrange(race) %>% 
  rename(`Race` = race, `Sample Count` = n) %>% 
  kable("latex", booktabs = TRUE, caption = "Sample counts by Race") %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

```{r fig-rna-pie-chart, fig.cap = "Detected RNA Types in iPSC Samples Analzyed using RNA Seq. The number of genes detected in each RNA category from RNA-seq data were counted and plotted as a pie chart (left). Small categories (< 2 %), such as rRNA, tRNA, and immunoglobulin genes, were combined into an 'Other' category. The expression of genes, using raw count per million data, was averaged by type of RNA and plotted as a pie chart (right).", fig.width=10, fig.height=5, echo=FALSE, warning=FALSE, message=FALSE}
plot_RNA_types_final + plot_RNA_expression_final 
```

```{r pca_sexage, echo=FALSE, warning=FALSE, message=FALSE}
#plot age_sex
pca_age_sex <- fviz_pca_biplot(
  pca_res,
  geom.ind = "point",                # samples as points
  habillage = pca_df$age_sex,           # color points by race
  addEllipses = TRUE,                # ellipses for race groups
  ellipse.level = 0.95,
  repel = TRUE,                      # avoid overlapping arrows
  geom.var = c("arrow", "text"),     # arrows + gene labels
  select.var = list(name = top_genes),  # only top variable genes
  col.var = "red",                    # arrows in red
  alpha.var = 0.7) +                   # slightly transparent arrows
labs(
  x = paste0("PC1 (", var_explained[1], "%)"),
  y = paste0("PC2 (", var_explained[2], "%)")) +
theme_classic(base_size = 14) +
theme(legend.position = "top")
```

```{r pca_race, echo=FALSE, warning=FALSE, message=FALSE}
#plot race
pca_race <- fviz_pca_biplot(
  pca_res,
  geom.ind = "point",                # samples as points
  habillage = pca_df$race,           # color points by race
  addEllipses = TRUE,                # ellipses for race groups
  ellipse.level = 0.95,
  repel = TRUE,                      # avoid overlapping arrows
  geom.var = c("arrow", "text"),     # arrows + gene labels
  select.var = list(name = top_genes),  # only top variable genes
  col.var = "red",                    # arrows in red
  alpha.var = 0.7) +               # slightly transparent arrows 
labs(
  x = paste0("PC1 (", var_explained[1], "%)"),
  y = paste0("PC2 (", var_explained[2], "%)")) +
theme_classic(base_size = 14) +
theme(legend.position = "top")
```

```{r fig_pca, fig.cap = "Principal Component Analysis (PCA) grouped by Donor's Age and Sex (top) and Race (bottom). PCA was performed using log2 gene expression across all samples, after removing genes with zero variance. PCA1 and PCA2 explain 10.5% and 6.4% of the variance in gene expression, respectively. Sample are colored by age-sex or race groups and ellipses represent 95% confidence intervals for each group. Two twenty genes contributing most to the variance, and thus most strongly influencing PCA1 and PCA2, are labeled as vectors.", fig.width=10, fig.height=14, echo=FALSE, warning=FALSE, message=FALSE}

grid.arrange(pca_age_sex, pca_race, ncol = 1)
```

```{r fig-pca-graph, fig.cap = "Principal Component Analysis (PCA) Grouped by Sex, Age, Age-Sex, and Race of Donors.  PCA was performed using log2 gene expression across all samples, after remvoing genes with zero variance. PCA1 and PCA2 explain 10.5% and 6.4% of the variance in gene expression, respectively. The samples were then colored by catagory: sex, age, age-sex, and race, to observe clustering of gene expression.", fig.width=11, fig.height=12, echo=FALSE, warning=FALSE, message=FALSE}

grid.arrange(PCA_sex, PCA_age, PCA_age_sex, PCA_race, ncol = 2, nrow = 2) 
```

```{r fig-heatmap-pluripotency, fig.cap = "Heatmap of Pluripotency Gene Expression. Rows correspond to genes and columns correspond to samples grouped by age-sex (top) and race (bottom).Colors indicate relative expression levels based on log2 count per million expression.  Hierarchical clustering was applied to the genes to highlight patterns of expression.", fig.width=8, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}

pluri_age_sex_hm <- pluri_agesex_results$heatmap
pluri_race_hm <- pluri_race_results$heatmap

pluri_g1 <- pluri_age_sex_hm$gtable
pluri_g2 <- pluri_race_hm$gtable

grid.arrange(pluri_g1, pluri_g2, ncol = 1) 
```

```{r fig-bargraph-pluripotency, fig.cap = "Expression of Pluripotency Genes by Age-Sex. Bar plots of log2 count per million gene expression by pluripotency gene. Bars represent mean expression ± standard error. Pairwise statistical significance was determined using Games-Howell post hoc tests following Welch’s ANOVA, and significance was indicated with asterisks:  p < 0.05 (\\texttt{*}), p < 0.01 (\\texttt{**}), p < 0.001 (\\texttt{***}).", fig.width=8, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
pluri_agesex_results$plot
```

```{r fig-bargraph-pluripotency-race, fig.cap = "Expression of Pluripotency Genes by Race. Bar plots of log2 count per million gene expression by pluripotency gene. Bars represent mean expression ± standard error. Pairwise statistical significance was determined using Games-Howell post hoc tests following Welch’s ANOVA, and significance was indicated with asterisks:  p < 0.05 (\\texttt{*}), p < 0.01 (\\texttt{**}), p < 0.001 (\\texttt{***})s", fig.width=8, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
pluri_race_results$plot
```

```{r pluri_agesex_stats, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
#pluripotency genes, age sex stats 

df_pluri_agesex <- pluri_agesex_results$data
 # Fit mixed model
#model the response variable, Expression, for the fixed effects, age and sex
# genes are treated as a random effect, all with different y intercepts,
# since expression levels vary by gene
lmm_pluri_agesex <- lmer(Expression ~ age_sex + (1 | `Gene name`), data = df_pluri_agesex)
summary_pluri_agesex <- summary(lmm_pluri_agesex)

  # type III ANOVA to test significance of fixed effects on expression
lmm_pluri_agesex_anova <- anova(lmm_pluri_agesex, type = 3)

# estimated marginal means (or least-squares means) with pairwise comparisons
lmm_pluri_agesex_emm <- emmeans(lmm_pluri_agesex, pairwise ~ age_sex)

#lmm summary
summary_pluri_agesex
#anova
lmm_pluri_agesex_anova
#means 
lmm_pluri_agesex_emm
#gene anova FDR corrected 
pluri_agesex_results$anova_fdr
#gene tueky corrected
pluri_agesex_results$Games_Howell_fdr
```

```{r pluri_race_stats, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
#pluripotency genes, race stats 

df_pluri_race <- pluri_race_results$data
 # Fit mixed model
#model the response variable, Expression, for the fixed effects, age and sex
# genes are treated as a random effect, all with different y intercepts,
# since expression levels vary by gene
lmm_pluri_race <- lmer(Expression ~ race + (1 | `Gene name`), data = df_pluri_race)
summary_pluri_race <- summary(lmm_pluri_race)
  # type III ANOVA to test significance of fixed effects on expression
lmm_pluri_race_anova <- anova(lmm_pluri_race, type = 3)

# estimated marginal means (or least-squares means) with pairwise comparisons
lmm_pluri_race_emm <- emmeans(lmm_pluri_race, pairwise ~ race)

#lmm summary
summary_pluri_race
#anova
lmm_pluri_race_anova
#means 
lmm_pluri_race_emm
#gene anova FDR corrected 
pluri_race_results$anova_fdr
#gene tueky corrected
pluri_race_results$Games_Howell_fdr
```

```{r pluri_lmm, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
summary_pluri_lmer <- lmer(Expression ~ age_decade * sex * race + 
                         (1 | `Gene name`), data = df_pluri_agesex)

summary(summary_pluri_lmer)
```

```{r fig-heatmap-diff, fig.cap = "Heatmap of Differentiation Gene Expression. Rows correspond to genes and columns correspond to samples grouped by age-sex (top) and race (bottom).Colors indicate relative expression levels based on log2 count per million expression.  Hierarchical clustering was applied to the genes to highlight patterns of expression.", fig.width=8, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
diff_age_sex_hm <- diff_agesex_results$heatmap
diff_race_hm <- diff_race_results$heatmap

diff_g1 <- diff_age_sex_hm$gtable
diff_g2 <- diff_race_hm$gtable

grid.arrange(diff_g1, diff_g2, ncol = 1) 
```

```{r fig-bargraph-diff, fig.cap = "Expression of Differentiation Genes by Age-Sex. Bar plots display log2 count per million gene expression by differentiation gene. Bars represent mean expression ± standard error. Pairwise statistical significance was determined using Games-Howell post hoc tests following Welch’s ANOVA, and significance was indicated with asterisks:  p < 0.05 (\\texttt{*}), p < 0.01 (\\texttt{**}), p < 0.001 (\\texttt{***}).", fig.width=8, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
diff_agesex_results$plot
```

```{r fig-bargraph-diff-race, fig.cap = "Expression of Differentiation Genes by Race. Bar plots display log2 count per million gene expression by differentiation gene. Bars represent mean expression ± standard error. Pairwise statistical significance was determined using Games-Howell post hoc tests following Welch’s ANOVA, and significance was indicated with asterisks:  p < 0.05 (\\texttt{*}), p < 0.01 (\\texttt{**}), p < 0.001 (\\texttt{***}).", fig.width=8, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
diff_race_results$plot
```

```{r diff_agesex_stats, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
#diff genes, age sex stats 

df_diff_agesex <- diff_agesex_results$data
 # Fit mixed model
#model the response variable, Expression, for the fixed effects, age and sex
# genes are treated as a random effect, all with different y intercepts,
# since expression levels vary by gene
lmm_diff_agesex <- lmer(Expression ~ age_sex + (1 | `Gene name`), data = df_diff_agesex)
summary_diff_agesex <- summary(lmm_diff_agesex)

  # type III ANOVA to test significance of fixed effects on expression
lmm_diff_agesex_anova <- anova(lmm_diff_agesex, type = 3)

# estimated marginal means (or least-squares means) with pairwise comparisons
lmm_diff_agesex_emm <- emmeans(lmm_diff_agesex, pairwise ~ age_sex)

#lmm summary
summary_diff_agesex
#anova
lmm_diff_agesex_anova
#means 
lmm_diff_agesex_emm
#gene anova FDR corrected 
diff_agesex_results$anova_fdr
#gene tueky corrected
diff_agesex_results$Games_Howell_fdr
```

```{r diff_race_stats, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
#diff genes, race stats 

df_diff_race <- diff_race_results$data
 # Fit mixed model
#model the response variable, Expression, for the fixed effects, age and sex
# genes are treated as a random effect, all with different y intercepts,
# since expression levels vary by gene
lmm_diff_race <- lmer(Expression ~ race + (1 | `Gene name`), data = df_diff_race)
summary_diff_race <- summary(lmm_diff_race)
  # type III ANOVA to test significance of fixed effects on expression
lmm_diff_race_anova <- anova(lmm_diff_race, type = 3)

# estimated marginal means (or least-squares means) with pairwise comparisons
lmm_diff_race_emm <- emmeans(lmm_diff_race, pairwise ~ race)

#lmm summary
summary_diff_race
#anova
lmm_diff_race_anova
#means 
lmm_diff_race_emm
```

```{r diff_lmm, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
summary_diff_lmer <- lmer(Expression ~ age_decade * sex * race + 
                         (1 | `Gene name`), data = df_diff_agesex)

summary(summary_diff_lmer)
```

```{r fig-heatmap-hk, fig.cap = "Heatmap of Housekeeping Gene Expression. Rows correspond to genes and columns correspond to samples grouped by age-sex (top) and race (bottom).Colors indicate relative expression levels based on log2 count per million expression.  Hierarchical clustering was applied to the genes to highlight patterns of expression.", fig.width=8, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
housekeeping_age_sex_hm <- housekeeping_agesex_results$heatmap
housekeeping_race_hm <- housekeeping_race_results$heatmap

housekeeping_g1 <- housekeeping_age_sex_hm$gtable
housekeeping_g2 <- housekeeping_race_hm$gtable

grid.arrange(housekeeping_g1, housekeeping_g2, ncol = 1) 
```

```{r fig-bargraph-hk, fig.cap = "Expression of Housekeeping Genes by Age-Sex. Bar plots display log2 count per million gene expression by housekeeping gene. Bars represent mean expression ± standard error. Pairwise statistical significance was determined using Games-Howell post hoc tests following Welch’s ANOVA, and significance was indicated with asterisks:  p < 0.05 (\\texttt{*}), p < 0.01 (\\texttt{**}), p < 0.001 (\\texttt{***}).", fig.width=8, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
housekeeping_agesex_results$plot
```

```{r fig-bargraph-hk-race, fig.cap = "Expression of Housekeeping Genes by Race. Bar plots display log2 count per million gene expression by housekeeping gene. Bars represent mean expression ± standard error. Pairwise statistical significance was determined using Games-Howell post hoc tests following Welch’s ANOVA, and significance was indicated with asterisks:  p < 0.05 (\\texttt{*}), p < 0.01 (\\texttt{**}), p < 0.001 (\\texttt{***}).", fig.width=8, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}
housekeeping_race_results$plot
```

```{r housekeeping_agesex_stats, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
#hk genes, age sex stats 

df_housekeeping_agesex <- housekeeping_agesex_results$data
 # Fit mixed model
#model the response variable, Expression, for the fixed effects, age and sex
# genes are treated as a random effect, all with different y intercepts,
# since expression levels vary by gene
lmm_housekeeping_agesex <- lmer(Expression ~ age_sex + (1 | `Gene name`), data = df_housekeeping_agesex)
summary_housekeeping_agesex <- summary(lmm_housekeeping_agesex)
  # type III ANOVA to test significance of fixed effects on expression
lmm_housekeeping_agesex_anova <- anova(lmm_housekeeping_agesex, type = 3)

# estimated marginal means (or least-squares means) with pairwise comparisons
lmm_housekeeping_agesex_emm <- emmeans(lmm_housekeeping_agesex, pairwise ~ age_sex)

#lmm summary
summary_housekeeping_agesex
#anova
lmm_housekeeping_agesex_anova
#means 
lmm_housekeeping_agesex_emm
#gene anova FDR corrected 
housekeeping_agesex_results$anova_fdr
#gene tueky corrected
housekeeping_agesex_results$Games_Howell_fdr
```

```{r housekeeping_race_stats, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
#diff genes, race stats 

df_housekeeping_race <- housekeeping_race_results$data
 # Fit mixed model
#model the response variable, Expression, for the fixed effects, age and sex
# genes are treated as a random effect, all with different y intercepts,
# since expression levels vary by gene
lmm_housekeeping_race <- lmer(Expression ~ race + (1 | `Gene name`), data = df_housekeeping_race)
summary_housekeeping_race <- summary(lmm_housekeeping_race)
  # type III ANOVA to test significance of fixed effects on expression
lmm_housekeeping_race_anova <- anova(lmm_housekeeping_race, type = 3)

# estimated marginal means (or least-squares means) with pairwise comparisons
lmm_housekeeping_race_emm <- emmeans(lmm_housekeeping_race, pairwise ~ race)

#lmm summary
summary_housekeeping_race
#anova
lmm_housekeeping_race_anova
#means 
lmm_housekeeping_race_emm
#gene anova FDR corrected 
housekeeping_race_results$anova_fdr
#gene tueky corrected
housekeeping_race_results$Games_Howell_fdr
```

```{r housekeeping_lmm, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
summary_hk_lmer <- lmer(Expression ~ age_decade * sex * race + 
                         (1 | `Gene name`), data = df_housekeeping_agesex)

summary(summary_hk_lmer)
```

```{r residual_function, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
plot_lmm_residuals <- function(lmm, data, fixed_effect, extreme_sd = 2) {
  
  # Extract residuals
  residuals_df <- data %>%
    mutate(
      Residual = resid(lmm, type = "pearson"),
      Residual_SD = Residual / sd(Residual),
      Extreme = abs(Residual_SD) > extreme_sd)
  
  # Histogram of residuals
  p_hist <- ggplot(residuals_df, aes(x = Residual, fill = Extreme)) +
    geom_histogram(binwidth = 0.2, color = "black") +
    scale_fill_manual(values = c("grey80", "red")) +
    theme_bw() +
    labs(title = paste("Residuals from LMM:", fixed_effect),
         x = "Residual", y = "Count")
  
  # QQ plot of residuals
  p_qq <- ggplot(residuals_df, aes(sample = Residual)) +
    stat_qq(aes(color = Extreme)) +
    stat_qq_line() +
    scale_color_manual(values = c("black", "red")) +
    theme_bw() +
    labs(title = paste("QQ plot of LMM residuals:", fixed_effect),
         color = "Extreme Residual")
  
  # List extreme residuals
  extreme_residuals <- residuals_df %>%
    filter(Extreme) %>%
    select(`Gene name`, Sample, Residual) %>%
    arrange(desc(abs(Residual)))
  
  # Automatically display plots
  print(p_hist)
  print(p_qq)
    # Display extreme residuals dataframe
  print(extreme_residuals)
  
  #return data invisibly
  invisible(list(
    residuals_df = residuals_df,
    extreme_residuals = extreme_residuals))
}

```

```{r histograms_and_qqplot, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
#histograms and QQ plots

#pluripotency genes
#residuals_pluri_age_sex <- plot_lmm_residuals(
 # lmm = summary_pluri_lmer,
  #data = df_pluri_agesex,
  #fixed_effect = c("age_decade", "sex", 'race'),
  #extreme_sd = 2)

#differentiation genes
#residuals_diff_age_sex <- plot_lmm_residuals(
 # lmm = summary_diff_lmer,
#  data = df_diff_agesex,
 # fixed_effect = c("age_decade", "sex", 'race'),
#  extreme_sd = 2)

#housekeeping genes
#residuals_housekeeping_age_sex <- plot_lmm_residuals(
 # lmm = summary_hk_lmer,
  #data = df_housekeeping_agesex,
  #fixed_effect = c("age_decade", "sex", 'race'),
  #extreme_sd = 2)
```

# Discussion

### RNA Composition

The RNA Seq data is composed of a variety of different types of RNA, as
seen in Figure 1. Of the genes detected, 34.8% of read genes are mRNA
protein coding genes, whereas the remaining 65.2% are not protein coding
and are primary composed of noncoding RNAs and pseudogenes, which are
RNA sequences that look like protein coding genes but are not able to be
translated into a protein. Specifically, 25.6% of the detected genes
were pseuodegenes, 25.5% were long non coding RNA sequences and 8.5%
were small RNAs, which included snoRNA, snRNA, scaRNA, sRNA, and
ribozyme RNA. 5.7% were "Other" RNAs, which include ribosomal RNA,
immunoglobin genes, T-cell receptor genes, and mitocondrial RNAs.

The right pie chart, on the other hand, displays the cumulative
expression of these groups. For each gene in each sample, expression was
reported as counts per million by Carcamo-Orive *et al.*. This is
typically done by dividing the raw count of the gene by the total reads
in the sample and multiplying by one million. The total expression was
summed for each category, and this was graphed in Figure 2, which
displays that 91.7% of the expression is from protein coding genes. The
remaining 8.3% accounts for the expression of small RNAs, long non
coding RNA, pseuodgenes, and "Other" genes. This is supported by Zhu *et
al*., who identfied that 93.7% of the transcripts in various RNA-Seq
analysed tissue samples were protein coding, while 6.3% were non-coding
transcripts [@zhu2016].

### PCA

Principal component anaysis (PCA) was used to identify what genes
expressed the greatest variation in expression of the iPSC analyzed
samples and to visualize how demographic variables contribute to their
variation. Figure 2 displays two PCA biplots annotated with 95%
confidence ellipses for age-sex groups (top) and race groups (bottom).
While the axis represent a low percentage of variability, 10.5% for PCA1
and 6.4% for PCA2, this is expected when a large number of genes
(58,302), and thus a large number of variables, are included in the
analysis. The ellipses displayed substantial overlap for the age-sex
groups, but race based ellipses displayed some clustering of samples,
suggesting that race contributes more strongly to the overall variations
in gene expression of the samples, seen in Figure 2. The PCA biplots
were then plotted as scatterplots and colored by age, sex, age-sex, and
race to further view clustering in the dataset, as seen in Figure 3.
This interpretation was quantified using a one way analysis of means,
not assuming equal variance due to the variation in the number of
samples per group, seen in Table 1 and 2. When tested independently,
neither age nor sex was significantly associated with PC1 or PC2.
However, when individuals were grouped by age–sex categories, a
significant effect emerged (Welch’s ANOVA, PCA1 p = 0.199 and PCA2 p =
0.265), suggesting that specific combinations of age and sex capture
expression patterns not explained by either variable alone.
Additionally, PCA1 and PCA2 were strongly influenced by the donor's race
(Welch's ANOVA, p \< 0.001). This suggests that environmental factors or
ancestry linked biological factors are major contributors to
transcriptional variability in iPSC lines.

To better understand the drivers of the observed PCA patterns, the top
twenty genes with the greatest variability in expression across the iPSC
samples were identified and labeled. These genes were primarily related
to genes found on the sex chromosomes, such as male specific genes
encoded on the Y chromosome: RPS4Y1 (var = 16.973), DDX3Y (var =
13.869), TXLNGY (var = 10.610), KDM5D (var = 10.160), USP9Y (var =
9.251), EIF1AY (var = 7.341), UTY (var = 7.231), NLGN4Y (var = 6.110);
an X linked gene: RPS4XP22 (var = 5.800); and genes involved in X
chromosome inactivation: XIST (var = 14.584) and TSIX (var = 12.091).
Since the iPSC sample population was composed of a mix of female and
male samples, this variability is to be expected, and was observed by
Caramo-Orive *et al*. They found that genes with greater than 2%
contributions to variation by sex were located on the X or Y chromosome
[@carcamo-orive2017].

The remaining nine genes with high variability in gene expression in
iPSCs samples were associated with a varietly of cell functions. CER1
(var = 7.325) and NLRP2 (var = 6.402) are associated with embryonic
development and early differentiation, tumor suppressors MEG3 (var =
5.980) and LINC00261 (var = 5.554P), the transcription factors ZNF208
(car = 6.294), APOA1 (var = 6.096) involved in lipid transport, LARGE1
(var = 6.874), a gene involved in glycosylation in the Golgi apparatuss,
CST1 (var = 9.794), a cysteine protease inhibitor , and COL3A1 (var =
6.400) for collagen type III, an extracellular matrix protein.
Caramo-Orive *et al* investigated the top 200 genes with the most
variability by performing a key driver analyses. They also found that
LINC00261, CER1, and APOA2 (similar to APOA1) were key drivers of
variability in iPSC samples, in addition to four other genes, GATA4,
GATA6, EOMES, and FOXQ1 [@carcamo-orive2017]. These genes could be
futher investigated to determine how their expression influences iPSC
pluripotency and differentiation.

### Pluripotency, Differentiation, and Houskeeping Gene Expression

The expression of pluripotency genes, SOX2, OCT4 (POU5F1), KLF4, c-Myc
(MYC), TRA-1-81 and 1-60 (PODXL), NANOG, SSEA4 (FUT4), LIN28A, and
Alkaline Phosphatase (ALPL)., were analyzed after grouping by both
age-sex and race. These genes include transcription factors, SOX2
(SRY-box transciption factor 2), OCT4 (octamer-binding transcription
factor 4), KLF4 (kruppel-like factor 4), and NANOG (nanog homeobox),
involved in maintaining a pluripotent state. Additionally, the
expression of PSC cell surface markers, SSEA4 (stage-specific embryonic
antigen-4) and TRA-1-81/TRA-1-60 (tumor related antigen), an RNA binding
protein, LIN28A (lin-28 homolog A), and a highly expressed enzyme in
pluripotent cells, alkaline phosphatase, were analyzed. These genes have
been found to be highly expressed in pluripotent cells, such as iPSCs
[@takahashi2007].

To determine if the expression of these pluripotent genes varied across
conditions, a linear mixed model was used. The predicted expression was
modeled with fixed effects age decade, sex and race, with random
intercepts grouped by gene to account for baseline differences in gene
expression. In the equation, the observation of each gene is represented
as 'o' , each gene is represented as 'g', and 'u' represents that the
intercept is unique to each gene.

$$
\hat{Expression}_{og} = \beta_0 + \beta_1 ({age\_decade}_{og}) + \beta_2 ({sex}_{og} ) + \beta_3 ({race}_{og}) + u_{0g} + \varepsilon
$$

Fixed-effect significance was assessed using t-tests with
Satterthwaite-approximated degrees of freedom to evaluate differences in
gene expression relative to the reference groups (female, age 40s,
African American).

Residual variance was 0.188 and gene intercept variance was 9.82, which
suggests that the majority of the variance is between genes, not between
groups.

It was found that there was no significant difference between female and
male samples, but there was a 6.5% decrease in expression in 70s samples
($\beta$ = - 0.097, p \< 0.05). The 70s female group displayed a 6.5%
decrease in pluripotency expression ($\beta$ = - 0.097, p \< 0.05.),
whereas 70s male saw a 13% increase in pluripotency expression ($\beta$
= + 0.179, p \< 0.001). There was no observed difference between the
other age-sex groups and the reference group (40s female), suggesting
that iPSCs generated from donors in their 70s differ in their expression
of pluripotency genes when compared to donors between the ages of 40 and
70. These results are observed in the Figure 4 top heatmap, which
display higher expression in 70s males for several pluripotency genes
(c-MYC, KLF4, SSEA4 (FUT4), and OCT4 (POU5F1)), and lower expression in
several genes for 70s female (ALPL, LIN28A, SOX2, TRA-1-81/TRA-1-60
(PODXL), and c-MYC). When looking at race, iPSC samples from South Asian
donors displayed a 10% decrease in pluripotency gene expression ($\beta$
= -0.143, p \< 0.01), but there was no difference between the other
groups and the reference group, African Americans. This can be seen in
Figure 4, where pluripotency expression for South Asian displayed both
increased (SSEA4 (FUT4) and KLF4) and decreased expression (OCT4
(POU5F1), NANOG, LIN28A, SOX2, and ALPL) in the majority of pluripotency
genes.

To look at the effect of individual genes, Welch's one-way ANOVA,
assuming unequal variance, followed by Games-Howell post hoc tests with
Benjamini-Hochberg corrected p values were used. SSEA4 (FUT4) displayed
eleven significant pairwise comparisons across age-sex groups, LIN28A
showed nine, and NANOG, POU5F1, and SOX2 had five significant pairwise
comparisons, seen in Figure 5. KLF4 and ALPL showed the fewest
significant differences, with only one pairwise comparison reaching
significance, but all genes displayed significant differences in gene
expression. For race, ALPL and LIN28A displayed five significant
pairwise comparisons, and SOX2 displayed four, seen in Figure 6. Several
pluripotency genes, including FUT4, NANOG, TRA-1-81 and 1-60 (PODXL),
and OCT4 (POU5F1), did not show any significant pairwise differences
across age-sex groups, indicating relatively stable expression.

Overall, pluripotency gene expresseion showed a higher variation in
age-sex groups compared to race groups, suggesting that donor age has a
greater effect on pluripotency gene expression in iPSC samples than sex
and race.

### Differentiation Markers

The expression of nine genes, which are commonly used as markers of
differentiation, were analyzed. Three genes were selected for each of
the three germ layers. For ectoderm markers, PAX6 (paired box 6) and
Nestin (NES, neuroepithelial stem cell protein) were used as neural
progenitor/stem cell markers, and KRT14 (keratin 14) as an epithelial
marker. SOX17 (SRY-box transcription factor 17) was used as a general
endoderm marker, AFP (alpha-fetoprotein) for hepatic cells (immature
hepatocytes), and PDX1 (pancreatic and duodenal homeobox 1) as a
pancreatic marker. For mesoderm markers, CD34 (cluster of differentation
34) for hematopoitic cells, CD90 (THY1, THYmocyte differentiation
antigen 1) for mesenchymal cells, and NKX2-5 (NK2 homeobox 5) for
cardiac progenitors were used. The expression of these linage specific
markers is typically observed when populations of iPSCs have begun to
randomly differentiation, and can be used to indicate the
differentiation pathway of the cells.

The same model that was used with the pluripotency genes was used to
analyze how age, sex, and race influence the expression of
differentiation genes but substituting the pluripotency genes with
differentiation genes.

Residual variance was 1.096 and gene intercept variance was 13.168,
which again suggests that the majority of the variance is between genes,
not between groups, however, these variances are larger than the
variances observed with the pluripotency genes.

It was found that there was no difference in the expression of
differentiation genes by sex, but there were statically significant
differences across age and race groups. All three age groups, (50s, 60s,
and 70s) displays differences in gene expression when compared to the
reference group, 40s. iPSCs from donors in their 50s displays a 70.8%
decrease ($\beta$ = -1.748, p \< 0.001), iPSCs from donors in their 60s
displays a 70.6% decrease ($\beta$ = -1.743, p \< 0.001), and iPSCs from
donors in thier 70s displayed a 57.7% increase ($\beta$ = + 0.655, p \<
0.001) in the expression of differentiation genes. This was further
broken down by age-sex groups, in which 50s female had a 17% decrease
($\beta$ = -0.269, p \< 0.001), 50s male displayed a 24.5% decrease
($\beta$ = -0.398, p\< 0.001), and 60s female showed a 29.3% decrease,
($\beta$ = -0.500, p \< 0.001).

When looking at race, there was a 25.3% decrease in the expression of
differentiation genes in iPSC samples from Asian donors ($\beta$ =
-0.429, p \< 0.001), a 32.0% increase for Hispanic donors ($\beta$ = +
0.399, p \< 0.01), and an 80.0% increase for South Asian donors ($\beta$
= + 0.854, p \< 0.001). These differences can be seen by gene in the
heatmap in Figure 7, in which the genes NKX2-5, PDX1, AFP, CD24, PAX6,
and SOX17 showed high expression in South Asian samples, but low
expression in Asian samples.

To look at the effect of individual genes, Welch's one-way ANOVA,
assuming unequal variance, followed by Games-Howell post hoc tests with
Benjamini-Hochberg corrected p values were used. KRT14, NKX2-5, and CD90
(THY1) displayed no statistically significant pairwise comparisons for
both age-sex and race groups, seen in Figure 8 and 9, suggesting that
there is minimal variation in the expression of these genes across the
iPSC samples. The remaining genes displayed one to two significant
pairwise comparisons (with p values \< 0.05 and 0.01) for the age-sex
groups, whereas these same genes displayed one to eight pairwise
comparisons (with p values \< 0.05, \< 0.01, and 0.001) when samples
were grouped by race. SOX17 and CD34, in particular displayed the most
pairwise comparisons, at eight and six respectively. This suggests that
the expression of differentiation markers is more heavily influenced by
donor race rather than by age and sex.

### Housekeeping Genes

The expression of nine commonly used reference/housekeeping genes in
mammalian cultures, such as stem cells, were analyzed. GAPDH
(glyceraldehyde-3-phosphate dehydrogenase) is one of the most widely
used housekeeping genes since it is involved in glycolysis and is
consistently expressed across various cell types [@panina2020].

Other commonly used gene involved in metabolism include PGK1
(phosphoglycerate kinase 1), cytoskelaton genes, ACTB (actin beta), and
ribosomal genes, RPL36AL (ribosomal protein L36a Like), and RPL13A
(ribosomal protein L13a). Other commonly used housekeeping genes include
B2M (beta-2 microglobulin), which in volved in the formation of major
histocompatibility complex (MHC) molecules, HPRT1 (hypoxanthine-guanine
phosphoribosyltransferase 1) which is an enzyme involved in the purine
salvage pathway, TBP for the TATA-box binding protein, and UBC for
ubiquitin C.

Residual variance was 0.148 and gene intercept variance was 7.16, which
suggests that the majority of the variance is between genes, not between
groups. Additionally, this is the lowest variance observed between the
three gene groups.

There was no difference in housekeeping expression of age groups 50s,
60s, and 70s when compared to the reference groups 40s, and a minimal
increase in expression for male (p \< 0.1). When looking at age-sex
groups, there were some differences in housekeeping gene expression,
with 9.6% increase in males in their 40s ($\beta$ = + 0.1325, p \<
0.001), a 2.4% increase for females in their 50s ($\beta$ = + 0.03375, p
\< 0.01), and a 5.7% increase for females in their 60s ($\beta$ = +
0.08191, p \< 0.01). This can be seen in the top heatmap in Figure 10,
where 40s males displayed higher expression of genes RPL36AL, GAPDH,
RPL13A, and PGK1 and lower expression of ACTB and UBC, when compared to
the reference groups, 40s female, which displayed lower expression in
RPL36Al, GAPDH, RPL13A, and B2M and higher expression in ACTB and UBC.

When looking at race, iPSCs from Hispanic donors showed a 7.1% increase
in housekeeping gene expression ($\beta$ = 0.09717, p \< 0.05), but
there was no statistically significant differences between the other
groups when compared to the reference group, African American. When
looking at the bottom heatmap in Figure 10, Asian and Caucasian samples
displayed consistent average expression values around zero, suggesting
that there is minimal variation in expression between housekeeping genes
for these samples. African American and South Asian samples diaplayed
the more variation, with a decrease in the expression for PGK1, HPRT1,
B2M, and UBC and increased expression of TBP, ACTB, and GAPDH for
African American samples, and decreased expression of ACTB, GAPDH,
RPL13A, and RPL36AL and increased expression of PDK1 for South Asian
samples.

To look at the effect of individual genes, Welch's one-way ANOVA,
assuming unequal variance, followed by Games-Howell post hoc tests with
Benjamini-Hochberg corrected p values were used. ACTB and RPl36AL
displayed no pairwise comparisons when grouping by age-sex, B2M, GAPDH,
HPRT1, TBP, RPL13A, and UBC displayed two to five significant
comparisons, and PGK1 displayed ten significant comparisons. This
suggests that PGK1 expression varies significantly when grouping by
age-sex, however, there were no significant comparisons when grouping by
race. RPL36AL, TBP, and UBC also displayed no significant comparisons
when grouping by race. ACTB, B2M, GAPDH, HPRT1, and RPL13A displayed one
to four comparisons.

Across all nine genes, RPL36AL is the only gene that displayed no
statistically significant pairwise comparisons for both age-sex groups
and race, suggesting that the expression of this gene is not influenced
by age, sex, or race and is consistently expressed across all of the
anazlyed iPSC samples. Overall, it appears that age and sex are the main
drivers of variation in housekeeping gene expression of the groups
studied here.

This is supported by the study conducted by Panina *et al* who
investigated the expression of 70 commonly used housekeeping genes
during iPSC reprogramming. They found that some of the most commonly
used genes, GAPDH, ACTB, HPRT, and B2M, fluctuated more than three fold
[@panina2020]. They also round that ribsosomal genes, such as RPL13A,
were the most stable regarding their expression during the reprogramming
process. In this study RPL13A expression appears to fluctuate in iPSC
samples, however, RPL36AL does not appear to.

# Conclusion

The transcriptome of 317 iPSC samples, collected using RNA-Seq from a
variety of donors, were analyzed to characterize overall RNA
composition, identify key drivers of expression variability via
principal component analysis (PCA), and access variability in the
expression of specific pluripotency, differentiation, and reference
genes. While a variety of RNA types were detected, 91.7% of the total
expression was contributed by protein coding mRNAs. PCA and Welch's
one-way ANOVA revealed that donor age-sex groups and race were major
contributors to expression variability, with the top genes driving
sample seperation including sex-linked genes located on the X and Y
chromosomes.

Expression of commonly used pluripotency markers, including LIN28A and
SOX2, varied across age–sex and race groups. Similarly, differentiation
markers such as SOX17 and CD34 exhibited numerous significant pairwise
differences, particularly between racial groups. Even housekeeping
genes, which are typically assumed to be stably expressed, displayed
variability across age–sex and race groups, notably GAPDH, HPRT1, and
PGK1. These findings indicate that donor demographic characteristics can
influence the transcriptome of iPSCs, including genes often used as
references or markers. However, because group sizes were unequal, some
variability may reflect limited sample numbers, even when using Welch’s
ANOVA with Benjamini-Hochberg p-value correction. Together, these
results demonstrate that donor demographics, including age, sex, and
race, significantly influence global gene expression patterns in iPSCs,
affecting both highly variable and commonly used reference genes.

------------------------------------------------------------------------

# References
