---
title: "Analysis of iPSC Gene Expression (will be changed)"
subtitle: "Final Project"
author: "Trinity Minard"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
format:
  pdf:
    keep-tex: true
editor_options:
  markdown:
    wrap: 72
bibliography:
  - references.bib
  - packages.bib
csl: norsk-apa-manual.csl
---

# Introduction

The discovery of human induced pluripotent stem cells (iPSCs) by
Takahashi and Yamanaka in 2007 [@takahashi2007] has revolutionized the
regenerative medicine field. iPSCs are created by reprogramming somatic
cells into an embryonic like state using the Yamanaka transciption
factors: SOX2, OCT4, KLF4, and c-Myc). These cells have vast
capabilities due to their ability to model human development and
diseases *in vitro*, and their potential for patient specific
(autologous) and "off the shelf" (allogeneic) cell therapies
[@cerneckis2024].

With that being said, the characterization of iPSC lines is essential in
any experimental or manufacturing process to ensure consistency in
culturing. This includes analyzing the expression of pluripotency (stem
cell) markers, differentiation markers, and potential chromosomal
abnormalies, to access the population of the culture. This is commonly
done using techniques such as reverse transcriptase or real time
quantitative PCR (RT-qPCR) to analyze mRNA expression [@artyukhov2017],
or using an antibody based experiment to analyze protein expression
levels, such as Flow Cytometry or immunocytochemistry. When conducting
these experiments, specifically RT-qPCR, reference genes, also known as
housekeeping genes, are used to normalize the expression of other genes
in a sample [@artyukhov2017]. This allows the expression of genes to be
compared across different samples, since it is presumed that the
expression of the housekeeping gene is consistent across all of the
samples. In stem cell samples, however, it has been found that the
commonly used housekeeping genes, such as GAPDH, appear to fluctuate in
iPSC samples [@panina2020], making it difficult to achieve accurate
normalization data.

RNA Seq is a technique that sequences the entirety of RNA molecules in a
given sample, rather than using gene specific primers or probes as in
RT-qPCR and other techniques. This allows the entire transciptome to be
analyzed, providing a complete picture of the expression of genes in a
sample [@wang2009]. Additionally, gene expression is calculated by
dividing the raw counts of a gene by the total counts in a sample,
eliminating the need for a housekeeping gene. This allows for an
unbiased and quantitiative analysis of millions of genes in sample
[@wang2009]. Despite these advantages, RNA-Seq is far more time
consuming and expensive than RT-qPCR and other experiments, making it
non practical in a manufacturing setting. Therefore, utilizing RNA-Seq
to analyze iPSC samples could provide insight into housekeeping gene
expression, allowing researchers to select ideal reference genes for
their analysis.

A similar study, conducted by Carcamo-Orive *et al.,* investigated the
variability of gene expression across 317 iPSC lines generated from 101
individuals [@carcamo-orive2017] and created a publically avaible data
set of their raw RNA Seq data. They found that genetic variability in
the iPSC lines was influenced by a donor specific gene expression
pattern, specifically influenced by donor age, sex, ancestry, and
reprogramming variabilies. iPSCs can also be reprogrammed from a variety
of somatic cells, such as blood, skin fibroblasts, hair keratinocytes,
urine cells, mesencymal stem cells, and bone marrow [@poetsch2022].
While some somatic cells are easier to obtain than others, such as blood
cells and fibroblasts, iPSCs from these lines are associated with higher
genomic mutations due to the somatic cells' high turnover rate.
Additionally, iPSCs appear to have an epigenetic memory of their tissue
of origin, which has been found to influence their differentiation
[@poetsch2022]. All of these factors, and many others, could be
contributing to the variations in gene expression in iPSC samples.

This publicly available dataset from Carcamo-Orive *et al.* was used to
analyze the expression of RNA across iPSC samples, with a focus on the
expression of pluripotency markers, differentiation markers, and
housekeeping genes.

# Methods

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#create list of packages to cite
packages <- c("base", "tidyverse", "dplyr", "ggplot2", "scales")
#citing R and packages
knitr::write_bib(packages, "packages.bib")
```

### Data Cleaning

RNA-Seq counts per million data was acquired from Stemformatics
[@stemform], using the Carcamo-Orive (2017) dataset
[@carcamo-orive2017]. Genes in this study were referred to by their
Ensembl ID, so gene information was integrated from Emsembl database
[@ensembl]. To begin to clean the data, the Ensembl gene ID in both the
count per million data frame and Ensembl data frame were mutated, using
the dplyr package [@R-dplyr] from tidyverse [@R-tidyverse], to the name:
Gene stable ID. The Ensembl gene data frame was then filtered for the
genes in the count per million data set (reducing the number of genes
from 86,369 to 56,693).

*further cleaning of the count per million data will be added*

### Figure 1

A new column was then added to the Ensembl data frame called Category,
which uses the Gene type column to categorize the genes into the
following categories: Protein coding, long non coding RNA (lncRNA),
Small RNAs, Pseudogenes, and Other. A count of the number of genes in
each category was then performed by using group by Category and
summarizing a count. A fraction and percentage were then calculated by
adding a column to the data frame that divides the count by the sum of
count (i.e. total count for all categories). This summarized data was
later plotted as a pie chart, so modifications were made to the
summarized data, such as defining ymax, ymin, and label position, for
ease of plotting. Text labeled were also created, which include the
category and the percentage calculated above. Percentages over 10% where
then added to a subcategory, called large slice, so that large
categories could be labeled on the slices but small categories would be
labeled outside of the pie chart. The scales [@R-scales] package from
ggplot2 [@R-ggplot2] was used to create the pie chart, by plotting "" on
the x axis, fraction on the y axis, and fill by category. Geom_bar and
coord_polar were used to format the pie chart, using fraction for the
size and angle of the slices. Theme_void was then used to remove the
default plot elements, creating a white background. Plot.margin was then
set to 0 on all sides to remove white space around the plot. Labels on
large slices were added by using the subset, large slice categories, and
adding a label using the label position and label text defined above.
Repel labels were added to categories not in large slices, using the
same label position and label text.

*continue methods for following figures ...*

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#library packages
library(tidyverse)
library(dplyr)
library(scales)
library(ggrepel)

#import data files
raw_counts <- read_tsv("stemformatics_dataset_7274.samples.tsv")
cpm_data   <- read_tsv("stemformatics_dataset_7274.cpm.tsv")
ensembl_data <- read_csv("mart_export.txt")

# view data files
#head(raw_counts)
#head(cpm_data)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#Change column name from "0.001" to "Gene Stable ID"
cpm_data <- cpm_data %>%
  rename("Gene stable ID" = "...1")

#Create a data frame with only the gene info for the genes in this study
ensembl_filtered <- ensembl_data %>%
  filter(`Gene stable ID` %in% cpm_data$`Gene stable ID`)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Add a new column 'Category' based on 'Gene type'
ensembl_filtered <- ensembl_filtered %>%
  mutate(Category = case_when(
    `Gene type` %in% c("protein_coding") ~ "Protein coding",
    `Gene type` %in% c("lncRNA", "TEC") ~ "lncRNA",
    `Gene type` %in% c("miRNA", 
                          "snoRNA", 
                          "snRNA", 
                          "scaRNA", 
                          "sRNA", 
                          "vault_RNA", 
                          "ribozyme") ~ "Small RNAs",
    `Gene type` %in% c("pseudogene", 
                          "processed_pseudogene", 
                          "unprocessed_pseudogene",
                          "transcribed_unprocessed_pseudogene",
                          "transcribed_processed_pseudogene",
                          "unitary_pseudogene", 
                          "IG_C_pseudogene", 
                          "IG_V_pseudogene",
                          "IG_J_pseudogene", 
                          "TR_J_pseudogene", 
                          "TR_V_pseudogene",
                          "translated_processed_pseudogene",
                        "transcribed_unitary_pseudogene") ~ "Pseudogenes",
       `Gene type` %in% c("misc_RNA", 
                          "artifact", 
                          "Mt_tRNA", 
                          "IG_V_gene", 
                          "IG_J_gene", 
                          "IG_C_gene", 
                          "IG_D_gene",
                          "TR_J_gene", 
                          "TR_V_gene", 
                          "TR_D_gene", 
                          "TR_C_gene",
                          "rRNA", 
                          "Mt_rRNA", 
                          "rRNA_pseudogene") ~ "Other",
   TRUE ~ "Other"))

#summarize count of genes by RNA catagory
RNA_types_summary <- ensembl_filtered %>%
  group_by(Category) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(fraction = count / sum(count),
         ymax = cumsum(fraction),
         ymin = c(0, head(ymax, n=-1)),
         label_pos = (ymax + ymin)/2,
         label_text = paste0(Category, "\n", scales::percent(fraction)),
         large_slice = fraction > 0.1)
```

# Results

```{r fig-rna-type, fig.cap = "Detected RNA Types. The number of genes detected in each RNA category from RNA-seq data were counted and plotted as a pie chart. Small categories (< 2 %), such as rRNA, tRNA, and immunoglobulin genes, were combined into an 'Other' category.", fig.width=8, fig.height=8, echo=FALSE, warning=FALSE, message=FALSE}

# Pie chart of number of genes per RNA catagory
ggplot(RNA_types_summary, aes(x = "", y = fraction, fill = Category)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  theme_void() +
theme(
  legend.position = "none",
  plot.margin = margin(0, 0, 0, 0)) +
  # On-slice labels for large slices
  geom_text(
    data = subset(RNA_types_summary, large_slice),
    aes(y = label_pos, label = label_text),
    size = 7,
    fontface = "bold",
    color = "white") +
  # Repelled labels for small slices
 geom_text_repel(
    data = subset(RNA_types_summary, !large_slice),
    aes(x = 1, y = label_pos, label = label_text),  
    size = 7,
    fontface = "bold",
    nudge_x = 0.8,
    force = 0.5,
    segment.color = "grey60",
    segment.length = 0.2,
    show.legend = FALSE) +
  labs(fill = "Category")
```

Figure 2. Expression by RNA Type (similar to figure 1 but expression
based rather than count based)

```{r fig-rna-expression, fig.cap = "Expression by RNA Type.", fig.width=8, fig.height=8, echo=FALSE, warning=FALSE, message=FALSE}
#

```

Figure 3. Expression of Pluripotency Genes (SOX2, OCT4, KLF4, c-Myc,
TRA-181, TRA-1-60, NANOG, SSEA4)

```{r fig-pluripotency, fig.cap = "Expression of Pluripotency Genes", fig.width=8, fig.height=8, echo=FALSE, warning=FALSE, message=FALSE}

protein_coding_genes <- ensembl_filtered %>%
  filter(`Gene type` == "protein_coding",
         !is.na(`Gene name`))
```

Figure 4. Expression of Differentiation Genes (ectoderm: PAX6, NESTIN or
SOX1 (neural projenitor/stem cell markers), KRT14 (epithelial),
endoderm: SOX17, AFP (hepatic), PDX1 (pancreatic), mesoderm: CD34
(hematopoitic), CD90 (mesenchymal), NKX2.5 (cardiac progenitor))

Figure 5. Expression of Housekeeping Genes (some sort of general figure
or sorted by housekeeping type (i.e. ribosomal, metabolic, cytoskeleton,
etc.))

Figure 6. Expression of Housekeeping Genes (focusing on some genes,
either over time during reprogramming or across different conditions)

these figures may be split into more than the six described depending on
the data

figures will include heat maps, bar charts, and possibly volcano plots

# Discussion

# Conclusion

------------------------------------------------------------------------

# References
